---
title: "TESSERA: Tool for Estimating Spatial and Sample-level Effects via Regression Analysis"
author:
  - name: Florica Constantine
    affiliation: "University of California, Berkeley"
package: TESSERA
output: 
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{TESSERA: Tool for Estimating Spatial and Sample-level Effects via Regression Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, warning=FALSE}
# Libraries

# TESSERA
library(TESSERA)

# Data Processing and Objects
# BiocManager::install("SummarizedExperiment")
# BiocManager::install("SingleCellExperiment")
# BiocManager::install("SpatialExperiment")
# BiocManager::install("ExperimentHub")
# BiocManager::install("spatialLIBD")
library(SummarizedExperiment)
library(SingleCellExperiment)
library(SpatialExperiment)
library(ExperimentHub)
library(spatialLIBD)

# Data normalization
# BiocManager::install("scran")
library(scran)

# General (Numerical) Utilties
library(dplyr)
library(reshape2)
library(Matrix)
library(MatrixGenerics)

# Statistics
library(MASS)
library(fdrtool)

# Visualization
library(ggplot2)
```

## Introduction

`TESSERA` is a package for fitting multi-sample generalized spatial linear models. 
Here, multi-sample indicates that we have multiple areas, tissue samples, or regions with independent spatial correlation structures. 
By generalized linear model, we mean that we fit an overdispersed Poisson model, where counts $z_j$ are 
$$z_j \sim \text{Pois}\left(C_j \exp\left(x_j^\top \beta + \phi_j\right) \right),$$
for some library size (offset) $C_j$, covariates $x_j$, random effects $\phi_j$, and fixed effects $\beta$. 
The $\phi_j$ are spatially correlated, with different samples having their own spatial correlation structures. 
Of particular note is that there is a single set of fixed effects $\beta$. 

This package is motivated by recent advances in spatial transciptomics, wherein multi-sample or multi-tissue data is becoming more common. 
More details are available in our paper [LINK TO PAPER].


## Installation

The following code will install the latest release version of the `TESSERA` package from Bioconductor. 
Additional details are shown on the [Bioconductor](https://bioconductor.org/packages/TESSERA) page.

```{r install, eval=FALSE}
install.packages("BiocManager")
BiocManager::install("TESSERA")
```

Alternatively, `TESSERA` can be installed from github
```{r install_git, eval=FALSE}
install.packages("remotes")
remotes::install_github("floricaconstantine/TESSERA")
```

## Loading Data

In the examples below, we assume that the data are provided as a SpatialExperiment object (Bioconductor). 
If this is not the case, `TESSERA` includes functionality to take in raw count matrices and metadata data frames in the `prepData` function. 

In what follows, we load and process data to ready it for `TESSERA`. 

```{r load_data}
# Connect to ExperimentHub
ehub <- ExperimentHub::ExperimentHub()
# Download the full real data:
spe_all <- spatialLIBD::fetch_data(type = "spe", eh = ehub)

# Assign row names to be gene names
rownames(spe_all) <- SummarizedExperiment::rowData(spe_all)$gene_name
```

Of particular note is that the rownames of the coordinates in a SpatialExperiment object might not match those of the metadata; it is not necessary to add the coordinates to the metadata, but it is necessary for the rownames to match. 

```{r}
# Add the spatial coordinates into the metadata (convenience)
SummarizedExperiment::colData(spe_all) <- cbind(SummarizedExperiment::colData(spe_all), 
                                            SpatialExperiment::spatialCoords(spe_all))
# Make sure rownames match and are unique
rownames(SummarizedExperiment::colData(spe_all)) <- SummarizedExperiment::colData(spe_all)$key
rownames(SpatialExperiment::spatialCoords(spe_all)) <- SummarizedExperiment::colData(spe_all)$key
colnames(SingleCellExperiment::counts(spe_all)) <- SummarizedExperiment::colData(spe_all)$key
```

This data set contains 4 samples from each of 3 subjects for a total of 12 samples. We subset to only the 4 samples from subject ``Br8100'' in this vignette to achieve a faster run-time in our method demonstration. 

```{r}
# Specify column names of spatial clusters in colData(spe) 
cluster_col <- "layer_guess_reordered"
# Remove spots missing annotations
spe_all <- spe_all[, !is.na(spe_all[[cluster_col]])]
# Subset to one subject
spe <- spe_all[, colData(spe_all)$subject == "Br8100"]
```

For our analysis, we focus on the top genes as sorted by biological variance. 
We see that the variance decreases rapidly. 

```{r sort_data}
# Identify mitochondrial genes (using the 'gene_name' column)
is_mito <- grepl("^MT-", rowData(spe)$gene_name, ignore.case = TRUE)

# Subset the object to keep only non-mitochondrial genes
spe <- spe[!is_mito, ]

# Fit the mean-variance trend, blocking by sample_id
dec <- scran::modelGeneVar(spe, block = spe$sample_id)

# Select the top based on biological variance
top_hvgs <- scran::getTopHVGs(dec, n = dim(spe)[1])

# Order spe by HVGs (biological variance)
hvg_order <- order(dec$bio, decreasing = TRUE)
spe <- spe[hvg_order, ]

# 
p <- ggplot2::qplot(seq_along(dec$bio), sort(dec$bio, decreasing = TRUE))
p <- p + ggplot2::scale_y_log10()
p <- p + ggplot2::labs(x = "Gene Index", y = "HVG: Biological Variance")
p <- p + theme_bw()
p
```

We visualize cluster annotations for all 4 samples from subject ``Br8100''. 

```{r visualize_covariates}
# Create the spatial plot
p <- ggplot2::ggplot(
  data = as.data.frame(SummarizedExperiment::colData(spe)),
  mapping = ggplot2::aes(x = pxl_col_in_fullres, y = pxl_row_in_fullres, color = layer_guess_reordered)
)
p <- p + ggplot2::geom_point(size = 1)
p <- p + ggplot2::facet_wrap( ~ sample_id)
p <- p + ggplot2::scale_y_reverse()
p <- p + ggplot2::scale_color_manual(values = c("darkorange", "gold", "forestgreen", "turquoise", "cornflowerblue", "purple", "hotpink"))
p <- p + ggplot2::labs(color = "Cluster")
p <- p + ggplot2::guides(color = ggplot2::guide_legend(override.aes = list(size = 4)))
p <- p + ggplot2::theme_void()
p <- p + ggplot2::theme(
  text = ggplot2::element_text(size = 20),
  plot.title = ggplot2::element_text(hjust = 0.5, face = "bold")
)
p
```

```{r visualize_neighbors}
# Figure out how to determine what cells are neighbors
nb_data <- TESSERA::visualizeNeighborDistances(
  meta_data = SummarizedExperiment::colData(spe),
  sample_col = "sample_id",
  coord_data = SpatialExperiment::spatialCoords(spe)
)

# Per the plot, it looks like each cell has 6 neighbors
# So, we use the mid point of the average of the 6th/7th furthest neighbor distances
# as a threshold for determining neighbors

nb_thresh <- mean(colMeans(nb_data$mean_nb_dist[, 6:7]))

p <- nb_data$distances_plot
p <- p + ggplot2::geom_hline(yintercept = nb_thresh, color = "red")
p <- p + ggplot2::theme_bw()
p
```

```{r visualize_covariates}
# Create the spatial plot
gene_idx <- which(SummarizedExperiment::rowData(spe)$gene_name == "RORB")
df <- as.data.frame(
  cbind(SummarizedExperiment::colData(spe)[, c("sample_id", "pxl_row_in_fullres", "pxl_col_in_fullres")], 
        expr = SingleCellExperiment::counts(spe)[gene_idx, ]))

p <- ggplot2::ggplot(
  data = df,
  mapping = ggplot2::aes(x = pxl_col_in_fullres, y = pxl_row_in_fullres, color = expr)
)
p <- p + ggplot2::geom_point(size = 1)
p <- p + ggplot2::facet_wrap( ~ sample_id)
p <- p + ggplot2::scale_y_reverse()
p <- p + ggplot2::labs(color = "Count")
p <- p + ggplot2::guides(color = ggplot2::guide_legend(override.aes = list(size = 4)))
p <- p + ggplot2::theme_void()
p <- p + ggplot2::theme(
  text = ggplot2::element_text(size = 20),
  plot.title = ggplot2::element_text(hjust = 0.5, face = "bold")
)
p
```

```{r prep_data, message=FALSE, warning=FALSE}
# Prepare data for the TESSERA package
# Simple design: Look at annotated region in v. out
TESSERA_data <- prepDataSpatialExperiment(spDataObject = spe, 
                                          sample_col = "sample_id",
                                          design_formula = as.formula( ~ layer_guess_reordered + sample_id),
                                          model_type = "Leroux",
                                          expected_num_neighbors = 6
                                          )
```

## Running `TESSERA` on a single gene

We now run `TESSERA` on a single gene, in this case, the first marker gene. 
The obligatory inputs to `TESSERA` are the data object, the gene name (row/rowname in the count matrix), and the model type (covariance structure). 
The other inputs are parameters of the ECM algorithm, in particular, how many iterations to run, how to determine early stopping, etc. 
```{r run_TESSERA, message=FALSE, warning=FALSE}
# Which gene to run
gene_name <- "RORB"
# Run TESSERA for the top gene
TESSERA_out <- TESSERA::TESSERA_lattice(
  TESSERAData_obj = TESSERA_data,
  gene_name = gene_name,
  model_type = "Leroux",
  em_iters = 200,
  # How many ECM iterations to run (maximum)
  opt_iters = 5,
  # How many CM steps to run per E-Step
  em_min_iters = 15,
  # Min ECM iterations before early stopping
  em_tol = 1e-3,
  # Tolerance for early stopping
  em_stopping = "rel_loglike",
  # How to determine early stopping: Relative change in log likelihood
  verbose = FALSE
)

print(TESSERA_out$time)
print("Number of Iterations")
print(ncol(TESSERA_out$MSE_tracker))
```

```{r TESSERA_results}
# Print fitted parameters
print("Estimated Coefficients")
print(TESSERA_out$beta_hat)

# Inspect a summary of the model fitting results
print("Summary")
print(TESSERA_out$performanceSummary)
```

As a point of comparison, we fit a negative binomial (NB) GLM. 
We see that the error in the estimated counts is substantially greater that with TESSERA, and that the residual autocorrelations are also much higher. 
Moreover, in addition to the fit quality being much worse, the estimated coefficients differ dramatically from those fit by `TESSERA`. 

```{r NB_GLM_comparison}
# Run an NB GLM for comparison
z_data <- sapply(TESSERA_data$counts_list, function (x) {
  as.vector(x[gene_idx, ])
})
z_data <- Reduce(c, z_data)
X_data <- Reduce(rbind, TESSERA_data$X_list)
lib_data <- Reduce(c, sapply(TESSERA_data$library_size_list, as.vector))
# We need some sort of starting point for the GLM to converge
glm_out <- MASS::glm.nb(z_data ~ 0 + X_data + offset(log(lib_data)), 
                        init.theta = mean(z_data) / var(z_data))

glm_preds <- predict(glm_out, type = "response")
glm_resid <- resid(glm_out, type = "response")

# Print MSE
cat("NB GLM MSE", mean(glm_resid^2), "\n")

# Compute Moran's I of residuals
GLM_resid_MI <- c()
for (s_name in names(TESSERA_out$start_idx_list)) {
  GLM_resid_MI <- c(
    GLM_resid_MI,
    TESSERA::calc_moran(
      glm_resid[TESSERA_out$start_idx_list[s_name]:(TESSERA_out$start_idx_list[s_name]
                                                    + nrow(TESSERA_data$X_list[[s_name]]) - 1)],
      TESSERA_data$coords_list[[s_name]][, 1],
      TESSERA_data$coords_list[[s_name]][, 2]
    )[1]
  )
}
names(GLM_resid_MI) <- names(TESSERA_out$start_idx_list)
cat("NB GLM Residual Moran's I", GLM_resid_MI, "\n")

print(coef(glm_out))

summary(glm_out)
```

```{r combine_results}
# Create data to visualize model fitting
plot_data <- data.frame(
  residuals = TESSERA_out$residuals,
  predictions = TESSERA_out$predictions,
  eta_hat = TESSERA_out$eta_hat,
  theta_hat = TESSERA_out$theta_hat,
  counts = Reduce(c, sapply(TESSERA_data$counts_list, function (x) {
    as.vector(x[gene_idx, ])
  })),
  x = Reduce(rbind, TESSERA_data$coords_list)[, 1],
  y = Reduce(rbind, TESSERA_data$coords_list)[, 2],
  glm_predictions = glm_preds,
  glm_residuals = glm_resid
)

# Add in sample
plot_data$sample <- NA
for (s_name in names(TESSERA_out$start_idx_list)) {
  plot_data$sample[TESSERA_out$start_idx_list[s_name]:(TESSERA_out$start_idx_list[s_name] + nrow(TESSERA_data$X_list[[s_name]]) - 1)] <- s_name
}
```

We first plot the expression (log-transformed counts). 

```{r visualize_counts}
p <- ggplot2::ggplot(data = plot_data,
                     mapping = ggplot2::aes(
                       x = x,
                       y = y,
                       color = log2(1 + counts)
                     ))
p <- p + ggplot2::geom_point(size = 1)
p <- p + ggplot2::facet_wrap( ~ sample)
p <- p + ggplot2::theme_bw()
p
```

We next plot the residuals for TESSERA; we see that the residuals are flat. 

```{r TESSERA_resid_plot}
p <- ggplot2::ggplot(data = plot_data,
                     mapping = ggplot2::aes(x = x, y = y, color = residuals))
p <- p + ggplot2::geom_point(size = 1)
p <- p + ggplot2::facet_wrap( ~ sample)
p <- p + ggplot2::theme_bw()
p
```

We now plot the residuals from the NB GLM. 
The scale of the residuals is much higher than those from `TESSERA`. 
```{r glm_resid_plot}
p <- ggplot2::ggplot(data = plot_data,
                     mapping = ggplot2::aes(x = x, y = y, color = glm_residuals))
p <- p + ggplot2::geom_point(size = 1)
p <- p + ggplot2::facet_wrap( ~ sample)
p <- p + ggplot2::theme_bw()
p
```

For the purposes of this vignette, we have already run `TESSERA` on the top 3,000 genes. 
The scripts to do are in the `inst/examples` folder, where `vignette_load_process.R` runs the preprocessing, `vignette_run_all.R` runs `TESSERA`, and `vignette_aggregate_results.R` aggregates the results. 
We load the results here and look at the performance results. 

```{r load_vignette_data}
vignette_data <- readRDS(file.path("..", "inst", "extdata", "vignette_processed_output_data2.rds"))
performance_df <- vignette_data$performance_df
wald_df <- vignette_data$wald_df

head(performance_df)
```

## Hypothesis testing

Of particular interest, after fitting models to several genes, is differential expression. 
Often, this is accomplished via testing contrasts of an estimated fixed effects vector. 
Below, we construct a contrast matrix and then apply it to the results from `TESSERA`. 

```{r wald_stats, warning=FALSE, eval=FALSE}
# Contrast matrix
layer_pairs <- combn(7, 2)
contrast_matrix <- matrix(
  data = 0,
  ncol = ncol(TESSERA_data$X_list[[1]]),
  nrow = ncol(layer_pairs)
)

rownames_contrast <- rep("", ncol(layer_pairs))
for (idx in 1:ncol(layer_pairs)) {
  # This model has an intercept, so all effects are relative to layer 1
  if (1 == min(layer_pairs[, idx])) {
    contrast_matrix[idx, max(layer_pairs[, idx])] <- 1
  } else {
    contrast_matrix[idx, max(layer_pairs[, idx])] <- 1
    contrast_matrix[idx, min(layer_pairs[, idx])] <- -1
  }
  rownames_contrast[idx] <- paste0("L", max(layer_pairs[, idx]), "-", "L", min(layer_pairs[, idx]))
}
rownames(contrast_matrix) <- rownames_contrast
colnames(contrast_matrix) <- colnames(TESSERA_data$X_list[[1]])

# Subset to comparing Layer 4 and Layer 5
contrast_matrix[which(rownames(contrast_matrix) == "L5-L4"), ]
```

We can inspect the dataframe created in the aggregation script, from running `TESSERA` on the top 3,000 genes:
```{r}
head(wald_df)
```

For a generalized linear mixed model (GLMM), which `TESSERA` is fitting a type of, the null distribution of Wald test statistics is not known. 
For enough samples (not cells), it will approach a normal distribution or a $\chi_1^2$ (depending on if the statistics are squared). 
However, in the finite sample regime, GLMMs are prone to bias and biased standard errors. 

If we are testing 100s or more genes, there is a path forward: we might use Efron's idea of estimating the null distribution from the data. 
In our case, we believe that the square of the Wald statistics computed previously can be modeled as a scaled, non-central $\chi_1^2$ distribution. 
The functions below provide functionality to find the optimal parameter estimates. 
Essentially, we look for a threshold below which most of the statistics come from the null; for these statistics, the p-values should look uniformly distributed on $(0, 1)$. 
The plots below provide diagnostics for how the parameters and empirical null distribution changes as a function of the threshold. 

```{r empirical_null, warning=FALSE}
# Empirical Null Estimation
# Sweep over thresholds to find the optimal distribution parameters for the empirical null
wald_thresh_selection <- TESSERA::selectWaldStatisticThreshold(
  wald_stats = wald_df$wald_stat_t^2,
  quantile_spacing = 0.01,
  metric = "Raw_MSE"
)
cat(
  "(Scale, Shift)",
  wald_thresh_selection$chi2_params,
  "; Threshold",
  wald_thresh_selection$threshold,
  "\n"
)
```

After fitting a null distribution and obtaining p-values, we may perform testing for differential expression, in this case, between regions. 

```{r sig_genes}
PVAL_THRESH <- 0.05

# Compute p-values
wald_df$wald_pval <- TESSERA::scaledNonCentralChi2PValues(wald_stats = wald_df$wald_stat_t^2,
                                                          chi2_params = wald_thresh_selection$chi2_params)

wald_df$wald_pval_adj <- NA
for(ci in unique(wald_df$contrast_indices)){
  local_idx <- which(wald_df$contrast_indices == ci)
  wald_df$wald_pval_adj[local_idx] <- p.adjust(wald_df$wald_pval[local_idx], method="BH")
}
wald_df %>% dplyr::group_by(contrast_indices) %>% dplyr::summarise(nsig = sum(wald_pval_adj < PVAL_THRESH)) %>% as.data.frame()

# Adjust p-values
#wald_df$wald_pval_adj <- p.adjust(wald_df$wald_pval, method = "BH")

l5_l4_indices <- grep("L7-L6", rownames(wald_df))

wald_df_sub <- wald_df[l5_l4_indices, ]

wald_df_sub$wald_pval_adj <- p.adjust(wald_df_sub$wald_pval, method = "BH")

cat("Empirical Null, number of significant genes: ", sum(wald_df_sub$wald_pval_adj < PVAL_THRESH), "\n")

#cat("Significant genes: ", wald_df[which(wald_df$wald_pval_adj < PVAL_THRESH), "gene"], "\n")
```

```{r fdrtool_comparison}
locfdr_out <- fdrtool::fdrtool(wald_df$wald_stat_t, cutoff.method = "locfdr", plot = FALSE)
fndr_out <- fdrtool::fdrtool(wald_df$wald_stat_t, cutoff.method = "fndr", plot = FALSE)

l5_l4_indices <- grep("L7-L6", rownames(wald_df))

p_adj_loc <- p.adjust(locfdr_out$pval[l5_l4_indices], method="BH")
sum(p_adj_loc < PVAL_THRESH)

p_adj_fndr <- p.adjust(fndr_out$pval[l5_l4_indices], method="BH")
sum(p_adj_fndr < PVAL_THRESH)


wald_df_sub$wald_pval_adj <- p.adjust(wald_df_sub$wald_pval, method = "BH")

locfdr_out$pval

cat("Local FDR, number of significant genes: ", sum(locfdr_out$lfdr < PVAL_THRESH), "\n")

cat("FNDR, number of significant genes: ", sum(fndr_out$lfdr < PVAL_THRESH), "\n")

```

```{r fdrtool_comparison}

l5_l4_indices <- grep("L4-L3", rownames(wald_df))

wald_df_sub <- wald_df[l5_l4_indices, ]

locfdr_out <- fdrtool::fdrtool(wald_df_sub$wald_stat_t, cutoff.method = "locfdr", plot = FALSE)
fndr_out <- fdrtool::fdrtool(wald_df_sub$wald_stat_t, cutoff.method = "fndr", plot = FALSE)

cat("Local FDR, number of significant genes: ", sum(locfdr_out$lfdr < PVAL_THRESH), "\n")

cat("FNDR, number of significant genes: ", sum(fndr_out$lfdr < PVAL_THRESH), "\n")

```

