---
title: "TESSERA: Tool for Estimating Spatial and Sample-level Effects via Regression Analysis"
author:
  - name: Florica Constantine
    affiliation: "University of California, Berkeley"
package: TESSERA
output: 
  BiocStyle::html_document:
    toc_float: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{TESSERA: Tool for Estimating Spatial and Sample-level Effects via Regression Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# TODO: Add BibTeX
```

## Introduction

`TESSERA` is a method for fitting multi-sample generalized spatial linear models. 
In this context, "multi-sample" refers to datasets comprising multiple distinct areas or regions, such as independent tissue slices, each characterized by its own unique spatial correlation structure.

### The Statistical Model

`TESSERA` fits an overdispersed Poisson model. For an observation $j$ in sample $i$, the counts $z_{i, j}$ are modeled as:
$$z_{i, j} \sim \text{Pois}\left(C_{i, j} \exp\left(x_{i, j}^\top \beta + \phi_{i, j}\right) \right),$$
where

* $C_{i, j}$ is the offset (e.g., library size),
* $x_{i, j}$ represents the covariates,
* $\beta$ represents the fixed effects,
* $\phi_{i, j}$ represents the spatial random effects. 

The $\phi_{i, j}$ are spatially correlated within each sample, with different samples having their own spatial correlation structures. 

A defining feature of `TESSERA` is its ability to estimate a **single set of fixed effects** $\beta$ across the entire dataset, even while permitting each individual sample its own independent spatial correlation structure.
This provides a rigorous foundation for addressing differential expression across experimental conditions--—such as comparing diseased versus healthy tissue—--while simultaneously accounting for the spatial autocorrelation inherent within each sample.

### Supported Spatial Correlation Structures

`TESSERA` supports two primary classes of spatial correlation structures, categorized by how they handle spatial dependencies.

#### Lattice Models
These models are designed for data where spatial relationships are defined via measurement-to-measurement adjacency matrices. Supported structures include:

* Leroux
* CAR (Conditional Autoregressive)
* SAR (Simultaneous Autoregressive)

#### Sparse Nearest Neighbor Gaussian Process (spNNGP)

`TESSERA` uses the spNNGP framework to model dependencies via a sparsified kernel matrix derived from pairwise distances between measurements. Supported kernel functions include:

* Matern ("Mat")
* Exponential ("Exp")
* Gaussian ("Gau")
* Spherical ("Sph")
    
#### Implementation Recommendations

Selecting the appropriate structure depends on your data's geometry:

* **Grid-like or Regularly Sampled Data**: If adjacency can be clearly defined, we recommend the Leroux lattice model as generally the most robust choice.
* **Irregularly Sampled Data**: For data where measurements do not follow a strict grid and adjacency of measurements cannot be determined, the spNNGP model with either the Matern or Exponential kernel is recommended.

### Motivation: Spatial Transcriptomics

The development of this package is driven by the rapid growth of spatial transcriptomics, where multi-sample or multi-tissue datasets are increasingly common. 
In this context, samples often consist of independent tissue slices with disparate coordinate systems, varying cell-type distributions, and differing underlying tissue architectures. 
These factors often preclude the use of traditional spatial methods that assume a single, continuous coordinate plane.
`TESSERA` bridges this gap by allowing for sample-specific spatial modeling within a unified global inference framework. 
For more details on the underlying methodology, please see our `TESSERA` methodological manuscript [@constantine2026tessera].

## Installation

Install the latest release version of `TESSERA` from [Bioconductor](https://bioconductor.org/packages/TESSERA) using the following code:
```{r install, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("TESSERA")
```

Alternatively, `TESSERA` can be installed from [GitHub](https://github.com/floricaconstantine/TESSERA):
```{r install_git, eval=FALSE}
if (!require("remotes", quietly = TRUE))
    install.packages("remotes")
remotes::install_github("floricaconstantine/TESSERA")
```

## Input Data

`TESSERA` accepts either Bioconductor `SpatialExperiment` objects or standard R matrices as input. 
If you are not using a SpatialExperiment object, you can provide your data as a numeric count matrix, a numeric matrix of spatial coordinates, and an accompanying metadata dataframe.

For detailed specifications on these formats, please refer to the [Format Data for `TESSERA`](#format-data) section.

## Load and Inspect Data

To get started, load TESSERA along with the necessary data containers and utility packages:
```{r libraries, message=FALSE, warning=FALSE}
# TESSERA
library(TESSERA)

# Data Structures
library(SummarizedExperiment)
library(SingleCellExperiment)
library(SpatialExperiment)

# Data Manipulation and Matrix Utilities
library(dplyr)
library(reshape2)
library(Matrix)
library(MatrixGenerics)

# Statistics
library(MASS)
library(fdrtool)

# Visualization
library(ggplot2)
library(viridis)
```

We demonstrate the workflow using a pre-processed kidney dataset [@abedini2024single], which contains 12,511 genes across 37,143 cells. 
This dataset is stored as `SpatialExperiement` object.

```{r load_data, message=FALSE, warning=FALSE}

# TODO: Move this data somewhere

# Load Data
spe <- readRDS("../../spatial/TESSERA_manuscript/data/vignette_processing_code/kidney_data_spe.rds")
dim(spe)
```

The dataset consists of 14 samples across three experimental conditions.
```{r filter_data}
# Column containing sample IDs
sample_col <- "orig.ident"
# Column containing disease condition
condition_col <- "Group"

as.data.frame(SummarizedExperiment::colData(spe)) %>%
  dplyr::distinct(dplyr::pick(dplyr::all_of(c(sample_col, condition_col)))) %>%
  dplyr::count(dplyr::pick(dplyr::all_of(condition_col)))
```

We examine the number of cells per sample. Each sample contains several thousand cells.
```{r}
# Number of cells per sample
sort(table(SummarizedExperiment::colData(spe)[, sample_col]))
```

For compatibility with standard visualization tools, we append the spatial coordinates directly to the metadata.

**Note**: It is essential for the row names of the coordinates in a SpatialExperiment object to match the row names of the metadata and the column names of the count matrix; it is not necessary to add the coordinates to the metadata, but it is necessary for the row names to match. 
```{r}
# Add the spatial coordinates into the metadata (convenience)
SummarizedExperiment::colData(spe) <- cbind(SummarizedExperiment::colData(spe), 
                                            SpatialExperiment::spatialCoords(spe))
```

To focus our analysis on the most informative features, we sort the genes by raw count variance. 
In a typical workflow, you might subset the data object to the top $N$ genes (e.g., the top 3,000 or 5,000 genes) before fitting the model.
```{r sort_data, fig.height=4, fig.width=6}
# Calculate variance across all spots for each gene
rv <- MatrixGenerics::rowVars(SingleCellExperiment::counts(spe))
names(rv) <- rownames(spe)

# Order the entire object by raw count variance (highest to lowest)
spe <- spe[order(rv, decreasing = TRUE), ]
```

Finally, we visualize the spatial distribution of cell types across all 14 samples.
```{r visualize_covariates, out.width="100%", fig.height=8, fig.width=10}
# Extract metadata
plot_df <- as.data.frame(SummarizedExperiment::colData(spe))

# Cell type colors
set.seed(200)
P26 <- Polychrome::createPalette(26, c("#ff0000", "#00ff00", "#0000ff"))
names(P26) <- unique(plot_df$celltype)

# Create the spatial plot
p <- ggplot2::ggplot(
  data = plot_df,
  mapping = ggplot2::aes(x = pxl_col_in_fullres, y = pxl_row_in_fullres, color = celltype)
)
p <- p + ggplot2::geom_point(size = 0.25)
p <- p + ggplot2::scale_colour_manual(name = "Cell type", values = P26)
p <- p + ggplot2::facet_wrap( ~ orig.ident)
p <- p + ggplot2::scale_y_reverse()
p <- p + ggplot2::guides(color = ggplot2::guide_legend(override.aes = list(size = 4), ncol = 1))
p <- p + ggplot2::theme_void()
p <- p + ggplot2::theme(
  panel.background = ggplot2::element_rect(fill = "white", color = NA),
  text = ggplot2::element_text(size = 15),
  plot.title = ggplot2::element_text(hjust = 0.5, face = "bold")
)
p
```

## Construct Design Matrix

In this analysis, we account for three primary factors: cell type, experimental condition, and sample identity. 
Because gene expression patterns often vary across conditions in a cell-type-specific manner, our design explicitly models these interactions. 
This allows us to test for differential expression within specific cell types while simultaneously controlling for sample-level biological and technical variation.

To achieve this, we use an intercept-free, interaction-only design. 
This formulation treats every combination of condition and cell type as a distinct state, which simplifies the construction of contrasts (e.g., comparing DKD vs. Control within a specific cell type) for hypothesis testing (detailed in the [Hypothesis Testing](#hyp-test) section).

Additionally, we account for sample-specific effects representing the individual-level biological or technical variation.
Because each sample belongs to only one experimental condition, we must nest the sample ID within the condition, following a framework similar to the `DESeq2` multi-factor approach.

We first map metadata columns to experimental factors and specify reference levels for each.
This keeps the design matrix columns in a consistent order, making it much easier to compare specific cell types or conditions during the hypothesis testing step.
```{r design_mat}
# Define covariate columns in metadata
celltype_col <- "celltype"
sample_col <- "orig.ident"
group_col <- "Group"
coord_cols <- c("pxl_row_in_fullres", "pxl_col_in_fullres")
covariates_keep <- c(group_col, celltype_col) # Use when creating design formula

# Define reference levels (baselines) for each factor (celltype, sample, and condition)
ref_celltype <- "C_TAL"
ref_sample <- "HK2874_ST"
ref_group <- "Control"

# Re-level factors to ensure the model uses the correct baseline
SummarizedExperiment::colData(spe)[, sample_col] <- relevel(factor(SummarizedExperiment::colData(spe)[, sample_col]), ref = ref_sample)
SummarizedExperiment::colData(spe)[, celltype_col] <- relevel(factor(SummarizedExperiment::colData(spe)[, celltype_col]), ref = ref_celltype)
SummarizedExperiment::colData(spe)[, group_col] <- relevel(factor(SummarizedExperiment::colData(spe)[, group_col]), ref = ref_group)
```

Multi-sample studies often require controlling for "sample effects" while samples are nested within broader experimental conditions. 
To account for these effects without introducing multi-collinearity, we assign a nested ID (1, 2, 3...) to each sample within its respective group.
```{r}
# Extract unique sample IDs for each disease group
control_ids <- unique(SummarizedExperiment::colData(spe)[, sample_col][SummarizedExperiment::colData(spe)[, group_col]  == "Control"])
DKD_ids <- unique(SummarizedExperiment::colData(spe)[, sample_col][SummarizedExperiment::colData(spe)[, group_col]  == "DKD"])
HKD_ids <- unique(SummarizedExperiment::colData(spe)[, sample_col][SummarizedExperiment::colData(spe)[, group_col]  == "HKD"])

# Initialize a new ID vector
new_id <- rep(0, nrow(SummarizedExperiment::colData(spe)))

# Assign a numeric index to samples within each group
for (idx in 1:length(control_ids)) {
  new_id[SummarizedExperiment::colData(spe)[, sample_col] == control_ids[idx]] <- idx
}
for (idx in 1:length(DKD_ids)) {
  new_id[SummarizedExperiment::colData(spe)[, sample_col] == DKD_ids[idx]] <- idx
}
for (idx in 1:length(HKD_ids)) {
  new_id[SummarizedExperiment::colData(spe)[, sample_col] == HKD_ids[idx]] <- idx
}

# Store the nested ID back into the spatial experiment object
SummarizedExperiment::colData(spe)$nested_id <- factor(new_id)
new_nested_sample_col <- "nested_id"
```

The design matrix is constructed using the formula `~ 0 + Group:celltype + Group:nested_id`. 
The `0 +` term removes the global intercept, allowing the model to estimate absolute means for each interaction term, denoted by the `:` operator.

For example, rather than modeling how the 'DKD' group differs relative to a 'Control' baseline, `TESSERA` estimates the mean for each condition independently. 
This approach makes calculating differences between groups much more intuitive when defining contrasts for hypothesis testing.
```{r}
# Build the formula string: ~ 0 + Group:celltype + Group:nested_id
design_formula <- as.formula(paste0(" ~ 0 + ", 
                                    group_col, ":", celltype_col, " + ", 
                                    group_col, ":nested_id"))
print(design_formula)

# Create the design matrix
X_mat <- model.matrix(design_formula, SummarizedExperiment::colData(spe))
```

Raw design matrices often produce messy column names (e.g., GroupDKD:nested_id2). 
We rename these to reflect the actual sample IDs they represent for better interpretability in downstream analysis.
```{r}
# Associate each nested ID with its sample and condition
sample_mapping <- colData(spe) %>%
  as.data.frame() %>%
  distinct(pick(all_of(c(group_col, sample_col, "nested_id"))))

# Map nested IDs back to original sample names for clarity
for (i in seq_len(nrow(sample_mapping))) {
  old_name <- paste0(group_col, sample_mapping[[group_col]][i], ":nested_id", sample_mapping$nested_id[i])
  new_name <- paste0(group_col, sample_mapping[[group_col]][i], ":", sample_mapping[[sample_col]][i])
  colnames(X_mat) <- gsub(old_name, new_name, colnames(X_mat), fixed = TRUE)
}
```

We drop zero columns. 
This removes any columns that represent combinations of factors not present in the data.
```{r}
# Toss out zero (empty) columns
X_mat <- X_mat[, colSums(X_mat) != 0]
```

We ensure the matrix is full rank (i.e., that there is no multi-collinearity that would prevent the model from being solved).
```{r}
# Ensure the design matrix is full rank
if (Matrix::rankMatrix(X_mat)[1] != ncol(X_mat)) {
  stop("The design matrix is not full rank. Check for linear dependencies.")
}
```

We examine the dimensions of the resulting design matrix, where the rows represent the total number of observations and the columns represent the parameters, $\beta$, to be estimated. 
It serves as a quick check to ensure the number of model coefficients matches the sum of your cell-type/group combinations and nested samples.
```{r}
# Review the design matrix dimensions
dim(X_mat)
```

The columns of the final design matrix $X$ represent the specific parameters, $\beta$, to be estimated, with each column name corresponding to a unique cell type, group, or nested sample effect."
```{r}
# Examine the design matrix column names
head(colnames(X_mat))
```

## Format Data for `TESSERA` {#format-data}

`TESSERA` requires a specialized data structure to handle multi-sample with within-sample spatial correlation structures.
A key goal of the preparation step is computational efficiency: we pre-compute quantities that are shared across all genes to drastically reduce the time required for model fitting.

### Choosing a Spatial Random Effect Model

The data preparation requirements depend on which spatial random effect model you select. 
`TESSERA` supports two primary classes of models, which differ in how they define and compute spatial correlation.
Your choice determines the preparation steps required.

#### Lattice Models (Leroux, CAR, SAR)
Lattice models define spatial correlation based on a measurement-measurement adjacency matrix (i.e., identifying which cells are physically adjacent).

* **Model Varieties:** 
  - **Leroux** is our preferred default; it decomposes variance into spatially correlated and uncorrelated components and offers the fastest pre-computation.
  - **CAR and SAR** are also supported but are slightly more computationally demanding during the preparation phase.
* **Pre-computation:** These models require a one-time pre-computation of eigenvalues for specific sparse matrices. While this takes a few seconds per sample, it makes the subsequent gene-level model fitting faster.
* **Memory Note:** The pre-computation step computes eigenvalues for an $n \times n$ matrix (where $n$ is the number of cells per sample). For $n < 3000$, this typically uses $< 1$ GB of RAM.

#### Sparse Nearest-Neighbor Gaussian Process (spNNGP)
spNNGP models define spatial correlation as a function of pairwise distances between cells using a kernel function (e.g., Matern, Exponential, Gaussian, or Spherical).

* **The "Sparse" Advantage:** The "sparse nearest-neighbor" improvement over a traditional Gaussian Process is an approximation that enables greater scalability by discarding the correlation between far-away pairs of cells.
* **No Pre-computation:** spNNGP models do not require adjacency matrices or eigenvalue pre-computation, making the initial preparation faster.
* **Computational Trade-off:** However, the inference process is more intensive and may be slower than lattice models, especially when processing thousands of genes.

### Defining Adjacency Matrix for Lattice Models

If you choose a lattice model, you must define a neighborhood structure by identifying which cells are physically adjacent.
`TESSERA` can accept a pre-computed adjacency matrix.
However, if the adjacency matrix is unknown, we provide internal utilities to determine cell-cell adjacency based on the data’s spatial coordinates.

For structured grid-based datasets, such as next-generation sequencing datasets like Visium, you can simply provide the `expected_num_neighbors` argument (typically 6 for Visium due to its hexagonal grid). 
`TESSERA` then automatically determines the distance threshold below which we consider cells to be adjacent required to satisfy this neighbor count across the sample.
Ultimately, `expected_num_neighbors` serves to automate the discovery of a valid `D_THRESH` argument for structured data, while `D_THRESH` itself offers the manual control required for more complex or non-grid tissue architectures.

In particular, for imaging-based or irregularly spaced data, determining adjacency is often more nuanced. 
We provide the `visualizeNeighborDistances()` function as a diagnostic tool; it generates box plots of the distance to the $k$-th nearest neighbor for every cell. 
A sharp "jump" in distance typically indicates the boundary where cells are no longer physically adjacent. 
For example, a jump after the sixth neighbor suggests that the seventh closest cell is likely not a direct neighbor.

However, in highly irregular or "noisy" spatial layouts, these plots may become difficult to interpret. 
In such cases, you may need to manually inspect your spatial coordinates, or process associated images of your data if available, to determine a reasonable distance cutoff. 
This value can be passed directly to `prepData()` via the `D_THRESH` argument. 

```{r visualize_neighbors, out.width="100%", fig.height=4, fig.width=6}
# Inspect distances to identify the neighbor threshold
nb_data <- TESSERA::visualizeNeighborDistances(
  meta_data = SummarizedExperiment::colData(spe),
  sample_col = sample_col,
  coord_data = SpatialExperiment::spatialCoords(spe)
)

# Per the plot, we see a jump after the 6th neighbor.
# We use the midpoint of the 6th and 7th neighbor distances as our threshold.
nb_thresh <- mean(colMeans(nb_data$mean_nb_dist[, 6:7]))

p <- nb_data$distances_plot
p <- p + ggplot2::geom_hline(yintercept = nb_thresh, color = "red")
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = ggplot2::element_text(size = 15))
p
```

### The `prepData` function

The `prepData` function act as wrappers that convert your data into the structured format `TESSERA` expects. 
You can provide data either as a `SpatialExperiment` object or as raw matrices and dataframes.

#### Function Inputs

#### 1. Data Input (Choose One)
| Argument | Description |
| :--- | :--- |
| `spDataObject` | A `SpatialExperiment` object. If provided, counts, metadata, and coordinates are extracted automatically. |
| **OR** | |
| `count_matrix` | A $genes \times cells$ count matrix. Rownames must be variable names. |
| `meta_data` | Dataframe of covariates where `rownames(meta_data)` matches `colnames(count_matrix)`. |
| `coord_data` | Coordinates for observations. Required for `spNNGP` or when computing adjacency from scratch. |

#### 2. Statistical Design (Choose One)
| Argument | Description |
| :--- | :--- |
| `design_formula` | A standard R formula object (e.g., `~ 0 + Group:celltype`). |
| `design_mat` | A pre-constructed design matrix. Zero columns will be dropped automatically. |

#### 3. Spatial & Model Configuration

| Argument | Description |
| :--- | :--- |
| `sample_col` | String identifying the column in metadata used to split data by sample. |
| `model_type` | The spatial model to be used: `"Leroux"`, `"CAR"`, `"SAR"`, or `"spNNGP"`; passing in `"ALL"` prepares `TESSERA` for any of the four. |
| `adj_mat` | **(Lattice only)** A pre-computed sparse adjacency matrix. |
| `D_THRESH` | **(Lattice only)** Distance threshold for assigning neighbors. Required if `adj_mat` is not provided. |
| `expected_num_neighbors` | **(Lattice only)** The average neighbors per cell (e.g., 6 for Visium). Used to automate `D_THRESH` calculation; ignored if `D_THRESH` is supplied. |
| `k_search` | **(Lattice only)** The maximum number of neighbors to search for when forming adjacency matrices. |


The data preparation step takes approximately 2 minutes on a machine with 16GB of RAM.
We will load a pre-computed result object for this vignette.
```{r load_TESSERA_data}
TESSERA_data <- readRDS("../../spatial/TESSERA_manuscript/data/vignette_processing_code/TESSERA_data.rds")
```

The following code block illustrates the exact code used to perform the data performation step.
If you wish to run the code locally, simply change the chunk header to `eval=TRUE`.
```{r prep_data, eval=FALSE}
# Prepare data for the TESSERA package
TESSERA_data <- TESSERA::prepData(
  x = spe,
  sample_col = sample_col,
  design_mat = X_mat,
  model_type = "Leroux",
  expected_num_neighbors = 6
)
saveRDS(TESSERA_data, "../spatial/TESSERA_manuscript/data/vignette_processing_code/TESSERA_data.rds", compress="xz")
```

The function returns a `TESSERA_data` object: a structured list containing coordinates, covariates, counts, library sizes, and design matrices, all indexed by sample. For lattice models, this object also stores the pre-computed adjacency and degree matrices along with the eigenvalues required to accelerate the model-fitting process. This sample-wise organization ensures that `TESSERA` can efficiently iterate through the data and perform downstream calculations when performing statistical inference.

## Running `TESSERA`

`TESSERA` fits models on a single gene at a time.
This independence means there is no information sharing between genes during the fitting process, which allows for straightforward parallelization across genes.

To fit a model, you must provide three core inputs: the processed data object, the specific gene name, and the desired covariance structure (e.g., "Leroux").

`TESSERA` uses an Expectation-Conditional Maximization (ECM) algorithm to estimate model parameters. 
While the default settings are robust for most spatial transcriptomics datasets, you can fine-tune the convergence behavior using the following settings:

* **Iteration Limits:** `em_iters` sets the maximum number of allowed iterations. `em_min_iters` enforces a minimum number of steps to prevent the algorithm from exiting before the log-likelihood has stabilized.

* **Convergence Tolerance:** `em_tol` is the numerical threshold for stopping, and `em_stopping` defines the metric used (typically the relative change in log-likelihood). The model exits when the change between iterations falls below this tolerance.

* **Optimization Intensity:** `opt_iters` controls the number of internal Conditional Maximization (CM) steps performed during each E-step. Increasing this can reduce the total number of iterations required for convergence but increases the computational cost of each individual iteration.

Fitting these models can be computationally intensive even with all the implemented speed-ups, taking approximately 12 minutes for a single gene on a machine with 16GB of RAM.
We will load a pre-computed result object for this vignette

```{r load_TESSERA_out, message=FALSE, warning=FALSE}
TESSERA_out <- readRDS("../../spatial/TESSERA_manuscript/data/vignette_processing_code/TESSERA_out_ACTA2.rds")
```

The following code block illustrates the exact parameters used to generate the model results for the "ACTA2" gene.
If you wish to run the model locally, simply change the chunk header to `eval=TRUE`.

```{r run_TESSERA, eval=FALSE, message=FALSE, warning=FALSE}
# Which gene to run
gene_idx <- which(rownames(spe) == "ACTA2")
# Run TESSERA
TESSERA_out <- TESSERA::TESSERA_lattice(
  # Output of prepData function
  TESSERAData_obj = TESSERA_data,
  # Gene
  gene_name = rownames(spe)[gene_idx],
  # Which covariance structure to use
  model_type = "Leroux",
  # How many ECM iterations to run (maximum)
  em_iters = 200,
  # How many CM steps to run per E-Step
  opt_iters = 5,
  # Min ECM iterations before early stopping
  em_min_iters = 15,
  # Tolerance for early stopping
  em_tol = 1e-3,
  # How to determine early stopping: Relative change in log likelihood
  em_stopping = "rel_loglike",
  # Whether to print out parameter values and convergence information at each iteration
  verbose = FALSE
)

# Output the run-time
print(TESSERA_out$time)
```

Once the model has converged, the output object provides a comprehensive summary of the performance in the `performanceSummary` data frame, allowing for a sample-wise breakdown of the results. 

By inspecting the fitted spatial covariance parameters, we can quantify the strength of the spatial signal for the lattice models via the identifiable parameters `gamma_hat` or `tau2_hat`. 
While individual spNNGP model parameters are not identifiable, their estimates are provided in the relevant `nugget_hat`, `sill_hat`, `range_hat`, and `smoothness_hat` columns, with `kernel_type`pecifying the Gaussian process kernel used.

The MSE results, such as `MSE_counts_sample` relative to `Mean_counts2_sample`, facilitate the evaluation of count prediction accuracy on a per-sample basis. 

Finally, the Moran's I metrics, specifically `Moran_residuals`, act as a critical diagnostic to ensure that the spatial structure has been adequately captured, indicating that the model has successfully accounted for the spatial dependency in the data.

```{r}
# Inspect a summary of the model fitting results
head(TESSERA_out$performanceSummary, 2)
```

The estimated fixed effects are stored in the `beta_hat` field. 
Each entry corresponds to a column in your design matrix.
```{r}
# Print fitted parameters
head(TESSERA_out$beta_hat)
```

### Comparing against an NB GLM

To assess the impact of spatial modeling, it is useful to compare `TESSERA` results against a non-spatial baseline. 
In this section, we compare our results with a standard Negative Binomial (NB) Generalized Linear Model (GLM), which treats observations as independent.

Because `TESSERA` uses list-structured data to accommodate multiple samples efficiently, we must first reformat the data into the single vectors and matrices required by standard R modeling packages like MASS.
```{r NB_GLM_comparison}
# Concatenate count data for the specific gene into a single vector across all samples
# z_data <- sapply(TESSERA_data$counts_list, function (x) {
#   as.vector(x[gene_idx, ])
# })
# z_data <- Reduce(c, z_data)
gene_idx <- which(rownames(spe) == "ACTA2")
z_data <- unlist(lapply(TESSERA_data$counts_list, function(x) as.vector(x[gene_idx, ])))

# Combine design matrices and library sizes across all samples
X_data <- Reduce(rbind, TESSERA_data$X_list)
lib_data <- Reduce(c, sapply(TESSERA_data$library_size_list, as.vector))
```

We fit the NB GLM using the `MASS::glm.nb` function. To ensure a fair comparison, we include the same design matrix and an offset term to account for variation in library sizes (sequencing depth) across cells.

```{r}
# Fit the non-spatial baseline model
glm_out <- MASS::glm.nb(z_data ~ 0 + X_data + offset(log(lib_data)))

# Extract predictions and residuals
glm_preds <- predict(glm_out, type = "response")
glm_resid <- resid(glm_out, type = "response")
```

A fundamental diagnostic for determining whether a spatial model is necessary is checking the residuals for spatial patterns. 
If a non-spatial GLM is sufficient, its residuals should be randomly distributed. 
However, if significant spatial autocorrelation remains (measured by Moran's I), it indicates that the GLM is failing to capture structured spatial signals that `TESSERA` is designed to model.
```{r}
# Compute Moran's I of residuals for each sample
GLM_resid_MI <- c()

for (s_name in names(TESSERA_out$start_idx_list)) {
# Subset residuals for the current sample
  start_idx <- TESSERA_out$start_idx_list[s_name]
  end_idx <- start_idx + nrow(TESSERA_data$X_list[[s_name]]) - 1
  current_resid <- glm_resid[start_idx:end_idx]
  
  # Calculate Moran's I using TESSERA's utility function
  mi_val <- TESSERA::calc_moran(
    current_resid,
    TESSERA_data$coords_list[[s_name]][, 1],
    TESSERA_data$coords_list[[s_name]][, 2]
  )[1]
  
  GLM_resid_MI <- c(GLM_resid_MI, mi_val)
}
names(GLM_resid_MI) <- names(TESSERA_out$start_idx_list)
```

To visualize the differences, we create a unified plotting object. We plot the raw counts, `TESSERA` residuals, and NB GLM residuals on a shared scale to highlight how each model handles spatial dependency.
```{r combine_results}
# Create data to visualize model fitting
plot_data <- data.frame(
  residuals = TESSERA_out$residuals,
  predictions = TESSERA_out$predictions,
  eta_hat = TESSERA_out$eta_hat,
  theta_hat = TESSERA_out$theta_hat,
  counts = Reduce(c, sapply(TESSERA_data$counts_list, function (x) {
    as.vector(x[gene_idx, ])
  })),
  x = Reduce(rbind, TESSERA_data$coords_list)[, 1],
  y = Reduce(rbind, TESSERA_data$coords_list)[, 2],
  glm_predictions = glm_preds,
  glm_residuals = glm_resid
)

# Add in sample
plot_data$sample <- NA
for (s_name in names(TESSERA_out$start_idx_list)) {
  plot_data$sample[TESSERA_out$start_idx_list[s_name]:(TESSERA_out$start_idx_list[s_name] + nrow(TESSERA_data$X_list[[s_name]]) - 1)] <- s_name
}

# Force symmetric limits based on the absolute maximum value
# This ensures 0 is always the 'mid' color
max_abs_val <- max(abs(c(plot_data$residuals, plot_data$glm_residuals)), na.rm = TRUE)
sym_limits <- c(-max_abs_val, max_abs_val)

# Range across both sets of residuals
# shared_range <- range(c(plot_data$residuals, plot_data$glm_residuals), na.rm = TRUE)
```

First, we visualize the log-transformed counts of gene "ACTA2".

```{r visualize_counts, out.width="100%", fig.height=8, fig.width=10}
p <- ggplot2::ggplot(data = plot_data,
                     mapping = ggplot2::aes(
                       x = x,
                       y = y,
                       color = log2(1 + counts)
                     ))
p <- p + ggplot2::geom_point(size = 0.25)
p <- p + ggplot2::facet_wrap( ~ sample)
p <- p + viridis::scale_color_viridis(name = "ACTA2\nExpression")
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = ggplot2::element_text(size = 15))
p
```

The `TESSERA` residuals are generally "flat," meaning the spatial random effects have successfully accounted for the spatial structure in the data.

```{r TESSERA_resid_plot, out.width="100%", fig.height=8, fig.width=10}
p <- ggplot2::ggplot(data = plot_data,
                     mapping = ggplot2::aes(x = x, y = y, color = residuals))
p <- p + ggplot2::geom_point(size = 0.25)
p <- p + ggplot2::facet_wrap( ~ sample)
p <- p + ggplot2::scale_color_gradient2(
    low = "#39FF14",  
    mid = "black",    
    high = "#FF10F0",
    midpoint = 0,
    limits = sym_limits,
    # The pseudo_log transformation expands the range near 0 
    # and compresses the tails without removing them
    trans = scales::pseudo_log_trans(sigma = 1), 
    name = "TESSERA\nresiduals"
  )
#p <- p + viridis::scale_color_viridis(name = "TESSERA\nresiduals", limits = shared_range)
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = ggplot2::element_text(size = 15))
p
```

In contrast, the residuals from the NB GLM often exhibit structured patterns and a much higher magnitude. 
This clustering of errors demonstrates that a non-spatial model is not adequate in capturing significant biological or technical spatial dependency.
```{r glm_resid_plot, out.width="100%", fig.height=8, fig.width=10}
p <- ggplot2::ggplot(data = plot_data,
                     mapping = ggplot2::aes(x = x, y = y, color = glm_residuals))
p <- p + ggplot2::geom_point(size = 0.25)
p <- p + ggplot2::facet_wrap( ~ sample)
p <- p + ggplot2::scale_color_gradient2(
    low = "#39FF14",  
    mid = "black",    
    high = "#FF10F0",
    #low = "#0571b0",  # Professional Blue
    #mid = "grey92",    # Exact zero
    #high = "#ca0020", # Professional Red
    midpoint = 0,
    limits = sym_limits,
    # The pseudo_log transformation expands the range near 0 
    # and compresses the tails without removing them
    trans = scales::pseudo_log_trans(sigma = 1), 
    name = "NB GLM\nresiduals"
  )
#p <- p + viridis::scale_color_viridis(name = "NB GLM\nresiduals", limits = shared_range)
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = ggplot2::element_text(size = 15))
p
```

### Parallelize `TESSERA`

`TESSERA` can be run in parallel (or serial) to fit to multiple genes. 
However, `TESSERA` is computationally intensive if run on hundreds or thousands of genes as is typically for genomic data, so it is best to parallelize this process on a high performance computing cluster. 

The following code (not run) shows an example of how to use the futures package to parallelize `TESSERA` on 50 genes. 

```{r parallel_TESSERA, warning=FALSE, message=FALSE, eval=FALSE}
library(future)
library(future.apply)

# How many genes to run
gene_idx_list <- seq(50)
print(length(gene_idx_list))
gene_names <- rownames(spe)[gene_idx_list] 

# Setup the backend
n_workers <- min(length(gene_names), parallel::detectCores() - 2)
if (.Platform$OS.type == "unix") {
  future::plan(future::multicore, workers = n_workers)
} else {
  future::plan(future::multisession, workers = n_workers)
}

# Increase memory limit for globals (2GB)
options(future.globals.maxSize = 2000 * 1024^2) 

t0 <- Sys.time()

# 2. Run future_lapply
# future.chunk.size = 1 ensures workers don't "batch" these 13-minute tasks.
TESSERA_results <- future.apply::future_lapply(gene_names, function(g_name) {
  
  TESSERA::TESSERA_lattice(
    TESSERAData_obj = TESSERA_data,
    gene_name       = g_name,
    model_type      = "Leroux",
    em_iters        = 200,
    em_min_iters    = 15,
    em_tol          = 1e-3,
    em_stopping     = "rel_loglike",
    verbose         = FALSE
  )
  
}, future.seed = TRUE, future.chunk.size = 1)

t1 <- Sys.time()

# Clean up
future::plan(future::sequential)

cat("Parallel processing of", length(gene_names), "genes took:", 
    round(difftime(t1, t0, units = "min"), 2), "minutes.\n")
```

## Hypothesis Testing {#hyp-test}

After fitting models to multiple (usually hundreds or thousands) genes, we aim to identify which genes are differentially expressed. 
We define differential expression as a statistically significant difference between the entries of the fixed-effect vector, $\beta$, that correspond to distinct experimental groups (e.g., Control vs. DKD within the 'VSMC' cell type).

By explicitly modeling the spatial covariance structure, `TESSERA` has been shown to effectively control the Type I error rate, preventing the inflation of false positives that can occur when spatial autocorrelation is ignored. 
For a full discussion of the method's statistical properties and Type I error rate benchmarks, please refer to the `TESSERA` methodological manuscript [@constantine2026tessera].

For the purposes of this vignette, we have already run `TESSERA` on the top 3,000 genes by raw count variance and load the results.
```{r tessera_multi_gene_results}
# Also load results from TESSERA
TESSERA_all <- readRDS("../../spatial/TESSERA_manuscript/data/vignette_processing_code/TESSERA_results.rds")
```

The performance results are in the `perf_df` dataframe.
```{r load_vignette_data}
performance_df <- TESSERA_all$perf_df
```

The contrasts, contrasts values and associated standard errors, and Wald statistics are in the `wald_df` dataframe.
```{r}
wald_df <- TESSERA_all$wald_df
head(wald_df, 2)
```

### Creating contrasts

Often, this is accomplished via testing contrasts of an estimated fixed effects vector. 
The rows of the contrast matrix correspond to different contrasts, and the columns to the entries of the fixed effects vector. 
We assume that each contrast is tested separately and is of the form 
$$\sum_{k = 1}^p c_k \beta_k = 0,$$
where $\beta_k$ is the $k^{th}$ entry of the fixed effects vector $\beta$ and the $c_k$ are coefficients that encode the contrasts. 

#### DE between cell types

We also demonstrate how to construct a contrast matrix for testing differential expression between two cell types $c_1$ and $c_2$. 
The contrasts of interest are 
$$\frac{1}{3}\left(\beta_{c_1, \textit{Control}} + \beta_{c_1, \textit{DKD}} + \beta_{c_1, \textit{HKD}}\right) - \frac{1}{3}\left(\beta_{c_2, \textit{Control}} + \beta_{c_2, \textit{DKD}} + \beta_{c_2, \textit{HKD}}\right).$$
```{r contrast_mat_across_ct}
cell_type_list <- levels(SummarizedExperiment::colData(spe)[, celltype_col])

# Pairs of cell types
celltype_pairs <- combn(cell_type_list, 2)
# Colnames in design matrix
covariate_names <- colnames(X_mat)

# Loop over cell type pairs
contrast_list <- list()

for (ct_idx in 1:ncol(celltype_pairs)) {
  idx1 <- which(grepl(celltype_pairs[1, ct_idx], covariate_names))
  idx2 <- which(grepl(celltype_pairs[2, ct_idx], covariate_names))
  
  # Create contrast matrix
  tmp <- matrix(data = 0, nrow = 1, ncol = length(covariate_names))
  tmp[1, idx1] <- 1 / length(idx1)
  tmp[1, idx2] <- -1 / length(idx2)
  rownames(tmp) <- c(paste0(celltype_pairs[1, ct_idx], "-", celltype_pairs[2, ct_idx]))
  
  contrast_list[[1 + length(contrast_list)]] <- tmp
}

celltype_contrast_mat <- Reduce(rbind, contrast_list)
colnames(celltype_contrast_mat) <- covariate_names
head(celltype_contrast_mat[, 1:5], 2)
```

We view the results of the across-cell type contrasts on the "ACTA2" gene below.
```{r wald_stats_across, warning=FALSE}
one_gene_ct_wald <- TESSERA::waldTestStastics(TESSERA_out, celltype_contrast_mat)
one_gene_ct_wald <- dplyr::select(one_gene_ct_wald, -c("contrast_string"))
head(one_gene_ct_wald, 2)
```

#### DE within a cell type between conditions 

Below, we construct a contrast matrix for testing differential expression within a cell type between diseases. 
```{r contrast_mat_within_ct}
cell_type_list <- levels(SummarizedExperiment::colData(spe)[, celltype_col])
condition_list <- levels(SummarizedExperiment::colData(spe)[, group_col])

# Pairs of conditions
condition_pairs <- combn(condition_list, 2)
# Colnames in design matrix
covariate_names <- colnames(X_mat)

# Loop over cell types
contrast_list <- list()
for (ct in cell_type_list) {
  for (d_idx in 1:ncol(condition_pairs)) {
    idx1 <- which(covariate_names == paste0(group_col, condition_pairs[1, d_idx], ":", celltype_col, ct))
    idx2 <- which(covariate_names == paste0(group_col, condition_pairs[2, d_idx], ":", celltype_col, ct))
    
    # If something is missing, skip the contrast
    if ((0 == length(idx1)) || (0 == length(idx2))) {
      next
    }
    
    # Create contrast matrix
    tmp <- matrix(data = 0, nrow = 1, ncol = length(covariate_names))
    tmp[1, idx1] <- 1
    tmp[1, idx2] <- -1
    rownames(tmp) <- c(paste0(ct, ":", condition_pairs[1, d_idx], "-", condition_pairs[2, d_idx]))
    
    contrast_list[[1 + length(contrast_list)]] <- tmp
  }
}

contrast_mat <- Reduce(rbind, contrast_list)
colnames(contrast_mat) <- covariate_names
head(contrast_mat[, 1:5], 2)
```

We view the results of the within-cell type contrasts on the "ACTA2" gene below.
```{r wald_stats_within, warning=FALSE}
one_gene_wald <- TESSERA::waldTestStastics(TESSERA_out, contrast_mat)
one_gene_wald <- dplyr::select(one_gene_wald, -c("contrast_string"))
head(one_gene_wald, 2)
```


We can inspect the dataframe created in the aggregation script, from running `TESSERA` on the top 3,000 genes.
Specifically, we look for which genes are differentially expression between conditions within a cell type.
For cell type $c$, the contrasts of interest are 
$$
\beta_{\textit{c, DKD}} - \beta_{\textit{c, Control}}, \qquad 
\beta_{\textit{c, HKD}} - \beta_{\textit{c, Control}}, \qquad
\beta_{\textit{c, HKD}} - \beta_{\textit{c, DKD}}.
$$

```{r wald_df}
wald_df <- wald_df[!is.na(wald_df$wald_stat_t), ]
head(wald_df, 2)
```

### Thresholding

For a generalized linear mixed model (GLMM), which `TESSERA` is fitting a type of, the null distribution of Wald test statistics is not known. 
For enough samples (not cells), it will approach a normal distribution or a $\chi_1^2$ (depending on if the statistics are squared). 
However, in the finite sample regime, GLMMs are prone to bias and biased standard errors. 

If we are testing hundreds or thousands of genes, there is a path forward: we might use Efron's idea of estimating the null distribution from the data. 
In our case, we believe that the square of the Wald statistics computed previously can be modeled as a scaled, non-central $\chi_1^2$ distribution. 
The functions below provide functionality to find the optimal parameter estimates. 
Essentially, we look for a threshold below which most of the statistics come from the null; for these statistics, the p-values should look uniformly distributed on $(0, 1)$. 

```{r}
wald_thresh_selection <- readRDS("../../spatial/TESSERA_manuscript/data/vignette_processing_code/wald_thresh_selection.rds")
```

```{r empirical_null, eval=FALSE}
t0 <- Sys.time()
# Empirical Null Estimation
# Sweep over thresholds to find the optimal distribution parameters for the empirical null
wald_thresh_selection <- TESSERA::selectWaldStatisticThreshold(
  wald_stats = wald_df$wald_stat_t^2,
  quantile_spacing = 0.01,
  metric = "Raw_MSE"
)
t1 <- Sys.time()
cat("Threshold selection of", nrow(wald_df), "statistics took:", 
    round(difftime(t1, t0, units = "min"), 2), "minutes.\n")
```

We look at the estimated scale and shift parameters for the scaled, non-central $\chi_1^2$ distribution.

```{r}
wald_thresh_selection$chi2_params
```

We see the selected Wald statistic threshold.

```{r}
wald_thresh_selection$threshold
```

### Obtaining $p$-values

After fitting a null distribution and obtaining p-values, we may perform testing for differential expression, in this case, between conditions. 

#### TESSERA Empirical Null Estimation Procedure

```{r sig_genes}
PVAL_THRESH <- 0.05

# Compute p-values
wald_df$wald_pval <- TESSERA::scaledNonCentralChi2PValues(wald_stats = wald_df$wald_stat_t^2,
                                                          chi2_params = wald_thresh_selection$chi2_params)
# Adjust p-values
for (cd in unique(wald_df$contrast_description)) {
  wald_df$wald_pval_adj[wald_df$contrast_description == cd] <- p.adjust(wald_df$wald_pval[wald_df$contrast_description == cd], method = "BH")
}
```

We summarize the results by looking at the number of significant genes for each comparison. 
As a sanity check, we see that cell types with very few cells in the dataset, such as RBC (red blood cells), find the fewest number of DE genes. 
```{r n_sig_plot, out.width="100%", fig.height=8, fig.width=10}
n_sig_df <- data.frame(wald_df 
                       %>% dplyr::group_by(contrast_description) 
                       %>% dplyr::summarise(n_sig = sum(wald_pval_adj < PVAL_THRESH, na.rm = TRUE)))
n_sig_df$celltype <- sapply(n_sig_df$contrast_description, function (x) {strsplit(x, ":")[[1]][1]})
n_sig_df$comparison <- sapply(n_sig_df$contrast_description, function (x) {strsplit(x, ":")[[1]][2]})

p <- ggplot2::ggplot(data = n_sig_df, mapping = ggplot2::aes(x = celltype, y = n_sig, fill = comparison))
p <- p + ggplot2::geom_col(position = ggplot2::position_dodge(preserve = "single"))
p <- p + ggplot2::scale_fill_manual(values = c("cornflowerblue", "goldenrod", "coral"))
p <- p + ggplot2::labs(x = "Cell Type", 
                       y = "Number of Significant Genes",
                       fill = "Comparison")
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(axis.text.x = ggplot2::element_text(angle = -90, hjust = 0))
p <- p + ggplot2::theme(text = ggplot2::element_text(size = 15))
p
```

#### Threshold Selection using the fdrtool FNDR Method

we can compare our findings against the False Non-Discovery Rate (FNDR) method implemented in the `fdrtool` package. 
This approach determines an optimal significance threshold by minimizing the FNDR, which can provide robust error control when the empirical null distribution is shifted or broader than expected.
```{r fdrtool_comparison}
fndr_out <- fdrtool::fdrtool(wald_df$wald_stat_t,
                             cutoff.method = "fndr",
                             plot = FALSE)

wald_df$wald_pval_fndr <- fndr_out$pval

# Adjust p-values
wald_df$wald_pval_adj_fndr <- NA
for (cd in unique(wald_df$contrast_description)) {
  wald_df$wald_pval_adj_fndr[wald_df$contrast_description == cd] <- p.adjust(wald_df$wald_pval_fndr[wald_df$contrast_description == cd], method = "BH")
}
```

Finally, we evaluate the robustness of our findings by comparing the number of significant genes identified via the FNDR method from `fdrtool` to those identified by `TESSERA`. As shown below, the FNDR-based adjustment yields a count of differentially expressed (DE) genes very similar to our model's results, indicating that `TESSERA` provides a consistent and reliable framework for discovery.
```{r}
sum(wald_df$wald_pval_adj_fndr < PVAL_THRESH, na.rm = TRUE)
```

We examine the number of significant genes using `TESSERA`'s empirical null estimation procedure.

```{r}
sum(wald_df$wald_pval_adj < PVAL_THRESH, na.rm = TRUE)
```

We can quantify the concordance between the two methods by identifying the intersection of the significant gene-contrast pairs. 
We find that all genes found to be significant by `TESSERA` are also found to be significant by `fdrtool`.
```{r}
# Create a unique identifier for each discovery (gene + contrast)
wald_df$discovery_id <- paste0(wald_df$gene, ":", wald_df$contrast_description)

# Identify significant discoveries for both methods
sig_tessera <- wald_df$discovery_id[wald_df$wald_pval_adj < PVAL_THRESH]
sig_fndr <- wald_df$discovery_id[wald_df$wald_pval_adj_fndr < PVAL_THRESH]

# Calculate the number of shared discoveries
n_tessera <- length(sig_tessera)
n_fndr <- length(sig_fndr)
n_intersection <- length(intersect(sig_tessera, sig_fndr))
n_intersection
```

The Jaccard Similarity between the two sets is also shown below. 
This confirms whether the methods are identifying the same biological signals rather than just finding similar total numbers of genes.

```{r}
# Calculate Jaccard Similarity (Intersection / Union)
jaccard <- n_intersection / length(union(sig_tessera, sig_fndr))
round(jaccard, 3)
```

## Session information

```{r sessioninfo}
sessionInfo()
```


