---
title: "TESSERA: Tool for Estimating Spatial and Sample-level Effects via Regression Analysis"
author:
  - name: Florica Constantine
    affiliation: "University of California, Berkeley"
package: TESSERA
output: 
  BiocStyle::html_document:
    toc_float: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{TESSERA: Tool for Estimating Spatial and Sample-level Effects via Regression Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = ">")
```

# Introduction

`TESSERA` is a method for fitting multi-sample spatial generalized linear mixed models (GLMM). 
In this context, "multi-sample" refers to datasets comprising multiple distinct areas or regions, such as independent tissue slices, each characterized by its own unique spatial covariance structure and coordinate system.

## Motivation: Spatial Transcriptomics

The development of `TESSERA` is driven by the rapid growth of spatial transcriptomics, where multi-sample or multi-tissue datasets are increasingly common. 
In this context, samples often consist of independent tissue slices with disparate coordinate systems, varying cell type distributions, and differing underlying tissue architectures. 

These factors often preclude the use of traditional spatial methods that rely on the assumption of a single, continuous coordinate plane.
`TESSERA` bridges this gap by allowing for sample-specific spatial modeling within a unified global inference framework. 
This allows investigators to ask different types of differential expression questions, both within and between samples. 
For more details on the underlying methodology, please see our `TESSERA` methodological manuscript [@constantine2026tessera].

## Limitations of Existing Methods

Traditional spatial statistical methods often rely on the assumption of a single, continuous coordinate plane. 
When applied to multi-sample studies, these frameworks typically break down because they cannot naturally bridge independent sample coordinates or accommodate experimental design hierarchies (e.g., nested technical or biological replicates across treatment groups). 
Investigators are frequently forced to either analyze samples in complete isolation—losing statistical power—or treat observations as spatially independent, which violates core assumptions and dramatically inflates False Discovery Rates (FDR).

# Method

`TESSERA` fits an overdispersed Poisson GLMM. For an observation $j$ in sample $i$, the counts $Z_{i, j}$ are modeled as:
$$Z_{i, j} \sim \text{Pois}\left(C_{i, j} \exp\left(\boldsymbol{X}_{i, j}^\top \boldsymbol{\beta} + \phi_{i, j}\right) \right),$$
where

* $C_{i, j}$ is the offset (e.g., library size),
* $\boldsymbol{X}_{i, j}$ represents the covariates,
* $\boldsymbol{\beta}$ represents the fixed effects,
* $\phi_{i, j}$ represents the spatial random effects. 

For each sample $i$, the spatial random effects follow:
$$\boldsymbol{\phi}_i \sim N(\mathbf{0}, \boldsymbol{\Sigma}_i),$$
where $\boldsymbol{\Sigma}_i$ is a sample-specific covariance matrix. 
This parameterization accounts for within-sample spatial autocorrelation while permitting different samples to have their own unique, independent spatial covariance structures.

A defining feature of `TESSERA` is its ability to estimate a **single set of fixed effects**, $\boldsymbol{\beta}$, across the entire dataset, even while permitting each individual sample its own independent spatial covariance structure.
This provides a rigorous foundation for addressing differential expression across experimental conditions--such as comparing diseased versus healthy tissue--while simultaneously accounting for the spatial autocorrelation inherent within each sample.

## Supported Spatial Covariance Structures

`TESSERA` supports two primary classes of spatial covariance structures, categorized by how they handle spatial dependencies.

### Lattice Models
These models are designed for data where spatial relationships are defined via measurement-to-measurement adjacency matrices. Supported structures include:

* Leroux
* CAR (Conditional Autoregressive)
* SAR (Simultaneous Autoregressive)

### Sparse Nearest-Neighbor Gaussian Process

`TESSERA` uses the sparse nearest-neighbor Gaussian process (spNNGP) framework to model dependencies via a sparsified kernel matrix derived from pairwise distances between measurements. Supported kernel functions include:

* Matern ("Mat")
* Exponential ("Exp")
* Gaussian ("Gau")
* Spherical ("Sph")
    
### Implementation Recommendations

Selecting the appropriate covariance model depends on the geometry of the dataset under consideration:

* **Grid-like or regularly-sampled data**: If adjacency can be clearly defined, we recommend the Leroux lattice model as generally the most robust choice given its flexible decomposition of spatial vs. non-spatial variance.
* **Irregularly-sampled data**: For data where measurements do not follow a strict grid and adjacency of measurements cannot be determined, the spNNGP model with either the Matern or Exponential kernel is recommended.

## Model Fitting via the ECM Algorithm

Parameter estimation is carried out using an Expectation-Conditional-Maximization (ECM) algorithm. The E-step computes the expected complete-data log-likelihood, while subsequent CM steps sequentially maximizes this log-likelihood by updating the fixed effects $\boldsymbol{\beta}$ and covariance parameters of the random effects.

## Hypothesis Testing & Empirical Nulls

Because finite-sample GLMMs are prone to structural asymptotic biases, the true null distribution of the Wald test statistics is typically unknown. 
`TESSERA` addresses this by modeling the Wald statistics as a scaled, non-central $\chi_1^2$ distribution, estimating an empirical null distribution directly from the data across thousands of parallelized per-gene models.

# Package Design

`TESSERA` isolates data engineering from model execution. For most standard workflows, you only need to interact with a few high-level functions and a unified data container.

## Object Classes and Data Flow

The package revolves around a specialized class structure designed to seamlessly handle multi-sample groupings:

* **Input Integration:** `TESSERA` acts as a downstream companion to the Bioconductor ecosystem. 
It natively ingests `SpatialExperiment` objects, automatically unpacking coordinates, counts, and spatial metadata.
* **The `TESSERA_data` Container:** The main preprocessing function, `prep_data()`, compiles these inputs into a structured, optimized list object. 
For lattice models, it pre-computes matrix eigenvalues and neighborhood weights once across the entire tissue layout, caching these static matrices so they do not have to be re-calculated during gene-by-gene iteration.

## Core Functions at a Glance

The table below outlines the primary functions you will encounter in a typical analysis pipeline, moving from exploratory analysis to downstream statistical inference.

| Function | Step | Purpose | Primary Inputs | Primary Outputs |
| :--- | :--- | :--- | :--- | :--- |
| `prep_data()` | **Data Prep** | Assembles inputs and pre-computes eigenvalues of spatial matrices. | Data matrix or `SpatialExperiment` object, Design matrix, model type | `TESSERA_data` object |
| `TESSERA_lattice()` / `TESSERA_spNNGP()` | **Fitting** | Fits the overdispersed GLMM model for a single gene. | `TESSERA_data`, gene name, optimization specs | Fitted model list object (`TESSERA_out`) |
| `calc_Wald_statistics()` | **Inference** | Conducts Wald tests against a user-defined contrast matrix. | `TESSERA_out`, contrast matrix | Data frame of test statistics |
| `select_Wald_threshold()` | **Inference** | Sweeps over statistics to fit the empirical null distribution. | Vector of Wald statistics | Empirical null parameter list |
| `calc_scaled_noncentral_chi2_pvalues()` | **Inference** | Computes empirical p-values using the fitted empirical null distribution. | Vector of Wald statistics, empirical null parameters | Vector of empirical p-values |

# Installation

Install the latest release version of `TESSERA` from [Bioconductor](https://bioconductor.org/packages/TESSERA) using the following code:
```{r install, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("TESSERA", dependencies = TRUE)
```

Alternatively, `TESSERA` can be installed from [GitHub](https://github.com/floricaconstantine/TESSERA):
```{r install_git, eval=FALSE}
if (!require("remotes", quietly = TRUE))
    install.packages("remotes")
remotes::install_github("floricaconstantine/TESSERA", dependencies = TRUE)
```

# Data Input & Exploration

## Input Data

`TESSERA` accepts either Bioconductor `SpatialExperiment` objects or standard R matrices as input. 
If you are not using a `SpatialExperiment` object, you can provide your data as a numeric count matrix, a numeric matrix of spatial coordinates, and an accompanying metadata data frame.

For detailed specifications on these formats, please refer to the [Pre-processing Data with `prep_data`](#prep-data) section.

## Load and Inspect Data

`TESSERA` utilizes the `Imports` namespace specification for its dependencies rather than `Depends`. 
This means that while all required packages are loaded internally to ensure `TESSERA` functions run correctly, their functions are not automatically attached to your global environment. 
If you wish to use non-`TESSERA` functions directly in your session, you will need to load those packages explicitly using `library()` or reference them using the double colon operator (e.g., `ggplot2::`). 
For clarity, this vignette explicitly loads all external packages required for the workflow below.
```{r libraries, message=FALSE, warning=FALSE}
# TESSERA
library(TESSERA)

# Data Downloading
library(BiocFileCache)

# Data Structures
library(SummarizedExperiment)
library(SingleCellExperiment)
library(SpatialExperiment)

# Data Manipulation and Matrix Utilities
library(dplyr)
library(reshape2)
library(Matrix)
library(MatrixGenerics)

# Statistics
library(MASS)
library(fdrtool)

# Visualization
library(ggplot2)
```

To keep the package lightweight while providing realistic, high-resolution spatial datasets, the data for this vignette is hosted externally. The following helper function uses `BiocFileCache` to download the necessary files from the `TESSERA_manuscript` repository and store them locally, ensuring that subsequent runs are fast and do not require an active internet connection.

```{r get_data}
# Define helper function to grab data files from the TESSERA_manuscript repository
get_my_data <- function(file_name) {
  # Define the URL (using the /raw/ path to handle Git LFS redirects)
  base_url <- "https://github.com/floricaconstantine/TESSERA_manuscript/raw/main/data/vignette_data/"
  target_url <- paste0(base_url, file_name)
  
  # Initialize the cache (ask = FALSE prevents interactive prompts during build)
  bfc <- BiocFileCache(ask = FALSE)
  
  # Check if the file is already in the cache, if not, download it
  # bfcrpath returns the LOCAL path to the cached file
  local_path <- bfcrpath(bfc, target_url)
  
  # Read the RDS from the local cached path
  return(readRDS(local_path))
}
```

We demonstrate the workflow using a pre-processed kidney dataset [@abedini2024single], which contains 12,511 genes across 37,143 cells. 
This dataset is stored as a `SpatialExperiment` object.
```{r load_data}
# Load Data
spe <- get_my_data("kidney_data_spe.rds")
dim(spe)
```

The dataset consists of 14 samples across three experimental conditions.
```{r filter_data}
# Column containing sample IDs
sample_col <- "orig.ident"
# Column containing disease condition
condition_col <- "Group"
# Column containing cell type
celltype_col <- "celltype"

as.data.frame(SummarizedExperiment::colData(spe)) %>%
  dplyr::distinct(dplyr::pick(dplyr::all_of(c(sample_col, condition_col)))) %>%
  dplyr::count(dplyr::pick(dplyr::all_of(condition_col)))
```

We examine the number of cells per sample. Each sample contains several thousand cells.
```{r num_cells_samp}
# Number of cells per sample
sort(table(SummarizedExperiment::colData(spe)[, sample_col]))
```

```{r plot_num_cells_samp}
# Convert table to a data frame
df_samp_counts <- as.data.frame(table(SummarizedExperiment::colData(spe)[, sample_col]))
colnames(df_samp_counts) <- c("sample", "count")

# Plot number of cells per sample
p <- ggplot2::ggplot(df_samp_counts, ggplot2::aes(x = reorder(sample, count), y = count)) 
p <- p + ggplot2::geom_bar(stat = "identity", fill = "steelblue") 
p <- p + ggplot2::theme_bw() 
p <- p + ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)) 
p <- p + ggplot2::labs(x = "Sample ID", y = "Number of cells")
p
```

We examine the number of cells per cell type across all cells and see that it varies widely by cell type.
```{r num_cells_celltype}
# Number of cells per cell type
sort(table(SummarizedExperiment::colData(spe)[, celltype_col]))
```

```{r plot_num_cells_ctyp}
# Convert table to a data frame
df_cty_counts <- as.data.frame(table(SummarizedExperiment::colData(spe)[, celltype_col]))
colnames(df_cty_counts) <- c("celltype", "count")

# Plot number of cells per sample
p <- ggplot2::ggplot(df_cty_counts, ggplot2::aes(x = reorder(celltype, count), y = count)) 
p <- p + ggplot2::geom_bar(stat = "identity", fill = "steelblue") 
p <- p + ggplot2::theme_bw() 
p <- p + ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)) 
p <- p + ggplot2::labs(x = "Cell type", y = "Number of cells")
p
```

For compatibility with standard visualization tools, we append the spatial coordinates directly to the metadata.

**Note**: It is essential for the row names of the coordinates in a SpatialExperiment object to match the row names of the metadata and the column names of the count matrix; it is not necessary to add the coordinates to the metadata, but it is necessary for the row names to match. 
```{r append_coords}
# Add the spatial coordinates into the metadata (convenience)
SummarizedExperiment::colData(spe) <- cbind(SummarizedExperiment::colData(spe), 
                                            SpatialExperiment::spatialCoords(spe))
```

To focus our analysis on the most informative features, we sort the genes by raw count variance. 
In a typical workflow, you might subset the data object to the top $N$ genes (e.g., the top 3,000 or 5,000 genes) before fitting the model.
```{r sort_data, fig.height=4, fig.width=6}
# Calculate variance across all spots for each gene
rv <- MatrixGenerics::rowVars(SingleCellExperiment::counts(spe))
names(rv) <- rownames(spe)

# Order the entire object by raw count variance (highest to lowest)
spe <- spe[order(rv, decreasing = TRUE), ]
```

Finally, we visualize the spatial distribution of cell types across all 14 samples.
```{r visualize_covariates, out.width="100%", fig.height=8, fig.width=10}
# Extract metadata
plot_df <- as.data.frame(SummarizedExperiment::colData(spe))

# Cell type colors
set.seed(200)
P26 <- Polychrome::createPalette(26, c("#ff0000", "#00ff00", "#0000ff"))
names(P26) <- unique(plot_df$celltype)

# Create the spatial plot
p <- ggplot2::ggplot(
  data = plot_df,
  mapping = ggplot2::aes(x = pxl_col_in_fullres, y = pxl_row_in_fullres, color = celltype)
)
p <- p + ggplot2::geom_point(size = 0.25)
p <- p + ggplot2::scale_colour_manual(name = "Cell type", values = P26)
p <- p + ggplot2::facet_wrap( ~ orig.ident)
p <- p + ggplot2::scale_y_reverse()
p <- p + ggplot2::guides(color = ggplot2::guide_legend(override.aes = list(size = 4), ncol = 1))
p <- p + ggplot2::theme_void()
p <- p + ggplot2::theme(
  panel.background = ggplot2::element_rect(fill = "white", color = NA),
  text = ggplot2::element_text(size = 15),
  plot.title = ggplot2::element_text(hjust = 0.5, face = "bold")
)
p
```

## Construct Design Matrix

In this analysis, we account for three primary factors: cell type, experimental condition, and sample identity. 
Because gene expression patterns often vary across conditions in a cell type-specific manner, our design explicitly models these interactions. 
This allows us to test for differential expression within specific cell types while simultaneously controlling for sample-level biological and technical variation.

To achieve this, we use an intercept-free, interaction-only design. 
This formulation treats every combination of condition and cell type as a distinct state, which simplifies the construction of contrasts (e.g., comparing DKD vs. Control within a specific cell type) for hypothesis testing (detailed in the [Hypothesis Testing](#hyp-test) section).

Additionally, we account for sample-specific effects representing the individual-level biological or technical variation.
Because each sample belongs to only one experimental condition, we must nest the sample ID within the condition, following a framework similar to the `r Biocpkg("DESeq2")` package [@love2014moderated] multi-factor approach.

We first map metadata columns to experimental factors and specify reference levels for each.
This keeps the design matrix columns in a consistent order, making it much easier to compare specific cell types or conditions during the hypothesis testing step.
```{r ref_lvls}
# Define covariate columns in metadata
celltype_col <- "celltype"
sample_col <- "orig.ident"
group_col <- "Group"
coord_cols <- c("pxl_row_in_fullres", "pxl_col_in_fullres")
covariates_keep <- c(group_col, celltype_col) # Use when creating design formula

# Define reference levels (baselines) for each factor (celltype, sample, and condition)
ref_celltype <- "C_TAL"
ref_sample <- "HK2874_ST"
ref_group <- "Control"

# Re-level factors to ensure the model uses the correct baseline
SummarizedExperiment::colData(spe)[, sample_col] <- relevel(factor(SummarizedExperiment::colData(spe)[, sample_col]), ref = ref_sample)
SummarizedExperiment::colData(spe)[, celltype_col] <- relevel(factor(SummarizedExperiment::colData(spe)[, celltype_col]), ref = ref_celltype)
SummarizedExperiment::colData(spe)[, group_col] <- relevel(factor(SummarizedExperiment::colData(spe)[, group_col]), ref = ref_group)
```

Multi-sample studies often require controlling for "sample effects" while samples are nested within broader experimental conditions 
To account for these effects without introducing multi-collinearity, we assign a nested ID (1, 2, 3...) to each sample within its respective condition.
```{r assign_nesting}
# Extract unique sample IDs for each experimental condition
control_ids <- unique(SummarizedExperiment::colData(spe)[, sample_col][SummarizedExperiment::colData(spe)[, group_col]  == "Control"])
DKD_ids <- unique(SummarizedExperiment::colData(spe)[, sample_col][SummarizedExperiment::colData(spe)[, group_col]  == "DKD"])
HKD_ids <- unique(SummarizedExperiment::colData(spe)[, sample_col][SummarizedExperiment::colData(spe)[, group_col]  == "HKD"])

# Initialize a new ID vector
new_id <- rep(0, nrow(SummarizedExperiment::colData(spe)))

# Assign a numeric index to samples within each condition
for (idx in 1:length(control_ids)) {
  new_id[SummarizedExperiment::colData(spe)[, sample_col] == control_ids[idx]] <- idx
}
for (idx in 1:length(DKD_ids)) {
  new_id[SummarizedExperiment::colData(spe)[, sample_col] == DKD_ids[idx]] <- idx
}
for (idx in 1:length(HKD_ids)) {
  new_id[SummarizedExperiment::colData(spe)[, sample_col] == HKD_ids[idx]] <- idx
}

# Store the nested ID back into the spatial experiment object
SummarizedExperiment::colData(spe)$nested_id <- factor(new_id)
new_nested_sample_col <- "nested_id"
```

The design matrix is constructed using the formula `~ 0 + Group:celltype + Group:nested_id`. 
The `0 +` term removes the global intercept, allowing the model to estimate absolute means for each interaction term, denoted by the `:` operator.

For example, rather than modeling how the 'DKD' condition differs relative to a 'Control' baseline, `TESSERA` estimates the mean for each condition independently. 
This approach makes calculating differences between conditions much more intuitive when defining contrasts for hypothesis testing.
```{r design_formula}
# Build the formula string: ~ 0 + Group:celltype + Group:nested_id
design_formula <- as.formula(paste0(" ~ 0 + ", 
                                    group_col, ":", celltype_col, " + ", 
                                    group_col, ":nested_id"))
print(design_formula)

# Create the design matrix
X_mat <- model.matrix(design_formula, SummarizedExperiment::colData(spe))
```

Raw design matrices often produce messy column names (e.g., GroupDKD:nested_id2). 
We rename these to reflect the actual sample IDs they represent for better interpretability in downstream analysis.
```{r rename_cols}
# Associate each nested ID with its sample and condition
sample_mapping <- colData(spe) %>%
  as.data.frame() %>%
  distinct(pick(all_of(c(group_col, sample_col, "nested_id"))))

# Map nested IDs back to original sample names for clarity
for (i in seq_len(nrow(sample_mapping))) {
  old_name <- paste0(group_col, sample_mapping[[group_col]][i], ":nested_id", sample_mapping$nested_id[i])
  new_name <- paste0(group_col, sample_mapping[[group_col]][i], ":", sample_mapping[[sample_col]][i])
  colnames(X_mat) <- gsub(old_name, new_name, colnames(X_mat), fixed = TRUE)
}
```

We drop zero columns. 
This removes any columns that represent combinations of factors not present in the data.
```{r drop_zeroes}
# Toss out zero (empty) columns
X_mat <- X_mat[, colSums(X_mat) != 0]
```

We ensure the matrix is full rank (i.e., that there is no multi-collinearity that would prevent the model from being fitted).
```{r full_rank}
# Ensure the design matrix is full rank
if (Matrix::rankMatrix(X_mat)[1] != ncol(X_mat)) {
  stop("The design matrix is not full rank. Check for linear dependencies.")
}
```

We examine the dimensions of the resulting design matrix, where the rows represent the observations and the columns represent the parameters, $\boldsymbol{\beta}$, to be estimated. 
It serves as a quick check to ensure the number of model coefficients matches the sum of the numbers of cell type/condition combinations and nested samples.
```{r dim_design_mat}
# Review the design matrix dimensions
dim(X_mat)
```

The columns of the final design matrix $X$ represent the specific parameters, $\boldsymbol{\beta}$, to be estimated, with each column name corresponding to a unique cell type, group, or nested sample effect.
```{r names_design_mat}
# Examine the design matrix column names
head(colnames(X_mat))
```

## Pre-processing Data with `prep_data` {#prep-data}

`TESSERA` employs a specialized data format designed to manage multiple samples while respecting the distinct spatial covariance structure inherent to each.
A key goal of the preparation step is computational efficiency: we pre-compute quantities that are shared across all genes to drastically reduce the time required for model fitting.

### The `prep_data` function

The `prep_data` function acts as a wrapper that converts your data into the structured format `TESSERA` expects. 
You can provide data either as a `SpatialExperiment` object or as raw matrices and data frames.

#### Function Inputs

#### 1. Data Input (Choose One)
| Argument | Description |
| :--- | :--- |
| `x` | A `SpatialExperiment` object. If provided, counts, metadata, and coordinates are extracted automatically. |
| **OR** | |
| `x` | A variables $\times$ measurements (e.g., genes $\times$ cells) count matrix. Rownames must be variable names. Colnames must be measurement IDs |
| `meta_data` | Data frame of covariates where `rownames(meta_data)` matches `colnames(count_matrix)`. |
| `coord_data` | Coordinates for observations. Required for `spNNGP` or when computing adjacency from scratch. |

#### 2. Statistical Design (Choose One)
| Argument | Description |
| :--- | :--- |
| `design_formula` | A standard R formula object (e.g., `~ 0 + Group:celltype`). |
| `design_mat` | A pre-constructed design matrix. Zero columns will be dropped automatically. |

#### 3. Spatial & Model Configuration

| Argument | Description |
| :--- | :--- |
| `sample_col` | String identifying the column in metadata used to split data by sample. |
| `model_type` | The spatial model to be used: `"Leroux"`, `"CAR"`, `"SAR"`, or `"spNNGP"`; passing in `"ALL"` prepares `TESSERA` for any of the four. |
| `adj_mat` | **(Lattice only)** A pre-computed sparse adjacency matrix. |
| `D_THRESH` | **(Lattice only)** Distance threshold for assigning neighbors. Required if `adj_mat` is not provided. |
| `expected_num_neighbors` | **(Lattice only)** The average neighbors per cell (e.g., 6 for Visium). Used to automate `D_THRESH` calculation; ignored if `D_THRESH` is supplied. |
| `k_search` | **(Lattice only)** The maximum number of neighbors to search for when forming adjacency matrices. |


The data preparation step takes approximately 2 minutes on an M1 MacBook with 16 GB of RAM.
We will load a pre-computed result object for this vignette.
```{r load_TESSERA_data}
# Output of prepData function
TESSERA_data <- get_my_data("TESSERA_prepData_out.rds")
```

The following code block illustrates the exact code used to perform the data preparation step.
If you wish to run the code locally, simply change the chunk header to `eval = TRUE`.
```{r prep_data, eval=FALSE}
# Prepare data for the TESSERA package
TESSERA_data <- TESSERA::prep_data(
  x = spe,
  sample_col = sample_col,
  design_mat = X_mat,
  model_type = "Leroux",
  expected_num_neighbors = 6
)
```

The function returns a `TESSERA_data` object: a structured list containing coordinates, covariates, counts, library sizes, and design matrices, all indexed by sample. For lattice models, this object also stores the pre-computed adjacency and degree matrices along with the eigenvalues required to accelerate the model-fitting process. This sample-wise organization ensures that `TESSERA` can efficiently iterate through the data and perform downstream calculations when performing statistical inference.

### Choosing a Spatial Random Effect Model

The data preparation requirements depend on the spatial random effect model selected. 
`TESSERA` supports two primary classes of models, which differ in how they define and infer spatial correlation.
The choice of model determines the preparation steps required.

#### Lattice Models (Leroux, CAR, SAR)
Lattice models define spatial correlation based on a measurement-measurement adjacency matrix (i.e., identifying which cells are physically adjacent).

* **Model Varieties:** 
  - **Leroux** is our preferred default; it decomposes variance into spatially correlated and uncorrelated components and offers the fastest pre-computation.
  - **CAR and SAR** are also supported but are slightly more computationally demanding during the preparation phase.
* **Pre-computation:** These models require a one-time pre-computation of eigenvalues for specific sparse matrices. While this takes a few seconds per sample, it makes the subsequent gene-level model fitting faster.
* **Memory Note:** The pre-computation step computes eigenvalues for an $n \times n$ matrix (where $n$ is the number of cells per sample). For $n < 3000$, this typically uses $< 1$ GB of RAM.

#### Sparse Nearest-Neighbor Gaussian Process (spNNGP)
spNNGP models define spatial correlation as a function of pairwise distances between cells using a kernel function (e.g., Matern, Exponential, Gaussian, or Spherical).

* **The "Sparse" Advantage:** The "sparse nearest-neighbor" improvement over a traditional Gaussian Process is an approximation that enables greater scalability by discarding the correlation between far-away pairs of cells.
* **No Pre-computation:** spNNGP models do not require adjacency matrices or eigenvalue pre-computation, making the initial preparation faster.
* **Computational Trade-off:** However, the inference process is more intensive and may be slower than lattice models, especially when processing thousands of genes.

### Defining Adjacency Matrix for Lattice Models

If you choose a lattice model, you must define a neighborhood structure by identifying which cells are physically adjacent.
`TESSERA` can accept a pre-computed adjacency matrix.
However, if the adjacency matrix is unknown, we provide internal utilities to determine cell-cell adjacency based on the data’s spatial coordinates.

For structured grid-based datasets, such as next-generation sequencing datasets from the Visium platform, you can simply provide the `expected_num_neighbors` argument (typically 6 for Visium due to its hexagonal grid). 
`TESSERA` then automatically determines the distance threshold required to satisfy the target neighbor count, defining any cells within this distance as adjacent.
Ultimately, `expected_num_neighbors` serves to automate the discovery of a valid `D_THRESH` argument for structured data, while `D_THRESH` itself offers the manual control required for more complex or non-grid tissue architectures.

In particular, for imaging-based or irregularly spaced data, determining adjacency is often more nuanced. 
We provide the `plot_neighbor_distances()` function as a diagnostic tool; it generates box-plots of the distance to the $k$-th nearest neighbor for every cell. 
A sharp "jump" in distance typically indicates the boundary where cells are no longer physically adjacent. 
For example, a jump after the sixth neighbor suggests that the seventh closest cell is likely not a direct neighbor.
```{r visualize_neighbors, out.width="100%", fig.height=4, fig.width=6}
# Inspect distances to identify the neighbor threshold
nb_data <- TESSERA::plot_neighbor_distances(
  meta_data = SummarizedExperiment::colData(spe),
  sample_col = sample_col,
  coord_data = SpatialExperiment::spatialCoords(spe)
)

# Per the plot, we see a jump after the 6th neighbor.
# We use the midpoint of the 6th and 7th neighbor distances as our threshold.
nb_thresh <- mean(colMeans(nb_data$mean_nb_dist[, 6:7]))

p <- nb_data$distances_plot
p <- p + ggplot2::geom_hline(yintercept = nb_thresh, color = "red")
p
```

However, in highly irregular or "noisy" spatial layouts, these plots may become difficult to interpret. 
In such cases, you may need to manually inspect your spatial coordinates, or process associated images of your data where available, to determine a reasonable distance threshold. 
This value can be passed directly to `prepData()` via the `D_THRESH` argument. 

# Model Fitting

## Running the ECM algorithm {#run-tessera}

`TESSERA` fits models on a single gene at a time.
This independence means there is no information sharing between genes during the fitting process and allows for straightforward parallelization across genes.

To fit a model, you must provide three core inputs: the `TESSERA_data` object generated by `prep_data` (see section [Pre-processing Data with `prep_data`](#prep-data)), the specific gene name, and the desired covariance structure (e.g., "Leroux").

`TESSERA` uses an Expectation-Conditional-Maximization (ECM) algorithm to estimate model parameters. 
While the default settings are robust for most spatial transcriptomics datasets, you can fine-tune the convergence behavior using the following settings:

* **Iteration Limits:** `em_iters` sets the maximum number of allowed iterations. `em_min_iters` enforces a minimum number of steps to prevent the algorithm from exiting before the log-likelihood has stabilized.

* **Convergence Tolerance:** `em_tol` is the numerical threshold for stopping, and `em_stopping` defines the metric used (typically the relative change in log-likelihood). The algorithm stops when the change between iterations falls below this tolerance.

* **Optimization Intensity:** `opt_iters` controls the number of internal Conditional-Maximization (CM) steps performed per E-step. Increasing this value performs additional parameter updates within each global ECM cycle. Because CM steps are much cheaper than E-steps, performing more CM steps increases the cost per iteration but minimizes the total number of global ECM cycles needed for convergence. 

Fitting the `TESSERA` model can be computationally intensive even with all the implemented speed-ups, taking approximately 12 minutes for a single gene on an M1 MacBook with 16 GB of RAM.
We will load a pre-computed result object for this vignette

```{r load_TESSERA_example}
# Output of TESSERA run with default settings for the "ACTA2" gene
TESSERA_out <- get_my_data("TESSERA_ACTA2_example.rds")
```

The following code block illustrates the exact parameters used to generate the model fitting results for the *ACTA2* gene.
If you wish to run the model locally, simply change the chunk header to `eval = TRUE`.
```{r run_TESSERA, eval=TRUE}
# Which gene to run
gene_idx <- which(rownames(spe) == "ACTA2")
# Run TESSERA
TESSERA_out <- TESSERA::TESSERA_lattice(
  # Output of prepData function
  TESSERAData_obj = TESSERA_data,
  # Gene
  gene_name = rownames(spe)[gene_idx],
  # Which covariance structure to use
  model_type = "Leroux",
  # How many ECM iterations to run (maximum)
  em_iters = 200,
  # How many CM steps to run per E-Step
  opt_iters = 5,
  # Min ECM iterations before early stopping
  em_min_iters = 15,
  # Tolerance for early stopping
  em_tol = 1e-3,
  # How to determine early stopping: Relative change in log likelihood
  em_stopping = "rel_loglike",
  # Whether to print out parameter values and convergence information at each iteration
  verbose = FALSE
)

# Output the run-time
print(TESSERA_out$time)
```

Once the algorithm has converged, the output object provides a comprehensive summary of the performance in the `performanceSummary` data frame, allowing for a sample-wise breakdown of the results. 

By inspecting the fitted spatial covariance parameters, we can quantify the strength of the spatial dependence for the lattice models via the identifiable parameters `gamma_hat` or `tau2_hat`. 
While individual spNNGP model parameters are not identifiable, their estimates are provided in the relevant `nugget_hat`, `sill_hat`, `range_hat`, and `smoothness_hat` columns, with `kernel_type` specifying the Gaussian process kernel used.

The MSE results, such as `MSE_counts_sample` relative to `Mean_counts2_sample`, facilitate the evaluation of fitted count accuracy on a per-sample basis. 

Finally, the Moran's $I$ metrics, specifically `Moran_residuals`, act as a critical diagnostic to ensure that the spatial structure has been adequately captured, indicating that the model has successfully accounted for the spatial dependency in the data.

```{r perf_summary}
# Inspect a summary of the model fitting results
head(TESSERA_out$performanceSummary, 2)
```

The estimated fixed effects are stored in the `beta_hat` field. 
Each entry corresponds to a column in your design matrix.
```{r beta_out}
# Print fitted parameters
head(TESSERA_out$beta_hat)
```

## Parallelization for multiple genes

For the purposes of this vignette, we have already run `TESSERA` on the top 3,000 genes (selected by raw count variance) as described in Section [Parallelize via HPC cluster](#parallelize-hpc).
The output data frames from each individual gene's run were row-bound to aggregate the results into a single combined data frame. We load this compiled data frame here to demonstrate downstream analysis and visualization.
```{r tessera_multi_gene_results}
# Load pre-computed results for the multi-gene analysis
TESSERA_all <- get_my_data("TESSERA_model_results.rds")
```

The `performance_df` data frame provides a sample-wise breakdown of model performance across the entire dataset. 
This object follows the same structure as the `performanceSummary` data frame described in the beginning of the [Run `TESSERA`](#run-tessera) section but contains rows for all gene-sample pairs.
```{r load_multi_data}
# Extract sample-level performance metrics for all 3,000 genes
performance_df <- TESSERA_all$perf_df
```

#### Parallelize via HPC cluster {#parallelize-hpc}

`TESSERA` can be run in parallel (or serial) to fit to multiple genes. 
However, `TESSERA` is computationally intensive if run on hundreds or thousands of genes as is typical for genomic data, so it is best to parallelize this process on a high-performance computing (HPC) cluster. 

The following code block illustrates the exact parameterization used to generate our aggregated results used for the remainder of this vignette. 
To process the full set of 3,000 genes, this model-fitting procedure was executed for each `gene_idx` (1 to 3,000) as independent batch jobs on a high-performance computing (HPC) cluster with the below parameterization.

```{r run_multi_TESSERA, eval=FALSE}
# Which gene to run
gene_idx <- 1
# Run TESSERA
TESSERA_out <- TESSERA::TESSERA_lattice(
  # Output of prepData function
  TESSERAData_obj = TESSERA_data,
  # Gene
  gene_name = rownames(spe)[gene_idx],
  # Which covariance structure to use
  model_type = "Leroux",
  # How many ECM iterations to run (maximum)
  em_iters = 200,
  # How many CM steps to run per E-Step
  opt_iters = 5,
  # Min ECM iterations before early stopping
  em_min_iters = 30,
  # Tolerance for early stopping
  em_tol = 1e-3,
  # How to determine early stopping: Relative change in log likelihood
  em_stopping = "rel_loglike",
  # Initialization of beta is done via Poisson GLM
  beta_init = "glm",
  # Initialization of gamma is based on the Moran's I of the log-transformed counts
  gamma_init = "moran",
  # Initialization of tau^2 is based on the variance of the log-transformed counts
  tau2_init = "var",
  # Whether to print out parameter values and convergence information at each iteration
  verbose = FALSE
)

# Output the run-time
print(TESSERA_out$time)
```

#### Parallelize via `futures` package {#parallelize-futures}

The following code (not run) shows an example of how to use the `futures` package to parallelize `TESSERA` on 50 genes. 

```{r parallel_TESSERA, eval=FALSE}
library(future)
library(future.apply)

# How many genes to run
gene_idx_list <- seq(50)
print(length(gene_idx_list))
gene_names <- rownames(spe)[gene_idx_list] 

# Setup the backend
n_workers <- min(length(gene_names), parallel::detectCores() - 2)
if (.Platform$OS.type == "unix") {
  future::plan(future::multicore, workers = n_workers)
} else {
  future::plan(future::multisession, workers = n_workers)
}

# Increase memory limit for globals (2GB)
options(future.globals.maxSize = 2000 * 1024^2) 

t0 <- Sys.time()

# Run future_lapply
# future.chunk.size = 1 ensures workers don't "batch" these 12-minute tasks.
TESSERA_results <- future.apply::future_lapply(gene_names, function(g_name) {
  
  TESSERA::TESSERA_lattice(
    TESSERAData_obj = TESSERA_data,
    gene_name       = g_name,
    model_type      = "Leroux",
    em_iters        = 200,
    em_min_iters    = 15,
    em_tol          = 1e-3,
    em_stopping     = "rel_loglike",
    verbose         = FALSE
  )
  
}, future.seed = TRUE, future.chunk.size = 1)

t1 <- Sys.time()

# Clean up
future::plan(future::sequential)

cat("Parallel processing of", length(gene_names), "genes took:", 
    round(difftime(t1, t0, units = "min"), 2), "minutes.\n")
```

# Hypothesis Testing {#hyp-test}

After fitting models to hundreds or thousands of genes, we aim to identify which among them are differentially expressed.
We define differential expression as a statistically significant difference between the entries of the fixed-effect vector, $\boldsymbol{\beta}$, that correspond to distinct experimental groups (e.g., Control vs. DKD within the *VSMC* cell type).

Unlike standard multi-sample methods that often treat observations as independent, `TESSERA` accounts for the inherent spatial correlation between measurements. 
By explicitly modeling this spatial covariance structure, the method effectively controls the Type I error rate. 
This prevents the inflation of false positives that commonly occurs when spatial autocorrelation is ignored--a frequent issue in non-spatial models that can lead to misleading biological conclusions.

For a comprehensive discussion of the model's statistical properties, including power and Type I error benchmarks, please refer to the `TESSERA` methodological manuscript [@constantine2026tessera].

The results of the hypothesis tests, including specific contrasts, estimated values, associated standard errors, and Wald statistics, are contained in the `wald_df` data frame.
Shown below are the results for between-condition comparisons performed within each cell type, a process elaborated upon in the [DE within a cell type between conditions](#de-within-celltype) section.

The aggregated results were obtained via the workflow described in Section [Parallelize via HPC cluster](#parallelize-hpc).
```{r multi_wald}
# Extract Wald test results for all gene-contrast pairs
wald_df <- TESSERA_all$wald_df
head(wald_df, 2)
```

## Creating Contrasts

Often, hypothesis testing is accomplished by testing contrasts of an estimated fixed-effects vector. 
The rows of the contrast matrix correspond to different contrasts, and the columns correspond to the entries of the fixed-effects vector. 
We assume that each contrast is tested separately and is of the form:
$$\sum_{k = 1}^p c_k \beta_k = 0,$$
where $\beta_k$ is the $k^{th}$ entry of the fixed effects vector $\beta$ and the $c_k$ are coefficients that encode the contrasts. 

We illustrate this approach below with two examples of contrast matrices designed to address distinct biological hypotheses: comparing expression across cell types and testing for condition-specific effects within a single cell type. 
These examples are not exhaustive; the framework is flexible, allowing for the construction of custom contrasts to suit a wide range of research objectives.

### DE between cell types

We demonstrate how to construct a contrast matrix for testing differential expression between two cell types $c_1$ and $c_2$. 
The contrasts of interest are 
$$\frac{1}{3}\left(\beta_{c_1, \textit{Control}} + \beta_{c_1, \textit{DKD}} + \beta_{c_1, \textit{HKD}}\right) - \frac{1}{3}\left(\beta_{c_2, \textit{Control}} + \beta_{c_2, \textit{DKD}} + \beta_{c_2, \textit{HKD}}\right).$$
First, we extract the cell type levels from the SpatialExperiment object and generate all possible pairwise combinations.
```{r between_pairwise}
# Get list of cell types from the SpatialExperiment object
cell_type_list <- levels(SummarizedExperiment::colData(spe)[, celltype_col])

# Generate all possible pairs of cell types
celltype_pairs <- combn(cell_type_list, 2)
# Column names in design matrix
covariate_names <- colnames(X_mat)
```

We iterate through each pair, identifying the corresponding columns in the design matrix. By assigning weights of $1/3$ (or $1/n$ depending on the number of conditions), we ensure we are testing the difference in average expression across all experimental groups.
```{r between_weights}
# Initialize an empty list to store individual contrast rows
contrast_list <- list()

# Loop over cell type pairs
for (ct_idx in 1:ncol(celltype_pairs)) {
  # Find indices in the design matrix for each cell type in the pair
  idx1 <- which(grepl(celltype_pairs[1, ct_idx], covariate_names))
  idx2 <- which(grepl(celltype_pairs[2, ct_idx], covariate_names))
  
  # Initialize a single-row matrix with zeros
  tmp <- matrix(data = 0, nrow = 1, ncol = length(covariate_names))

  # Assign weights to represent the average across conditions
  tmp[1, idx1] <- 1 / length(idx1)
  tmp[1, idx2] <- -1 / length(idx2)
  
  # Name the row for clear identification in the output
  rownames(tmp) <- c(paste0(celltype_pairs[1, ct_idx], "-", celltype_pairs[2, ct_idx]))
  contrast_list[[1 + length(contrast_list)]] <- tmp
}
```

Finally, we bind the individual rows together to create the full contrast matrix.
```{r between_bind}
# Combine the list of rows into a single contrast matrix
celltype_contrast_mat <- Reduce(rbind, contrast_list)
colnames(celltype_contrast_mat) <- covariate_names

# Preview the first few rows and columns
head(celltype_contrast_mat[, 1:6], 2)
```

We now view the results of the between-cell type contrasts for the gene *ACTA2* below.
```{r between_wald_stats}
# Compute Wald statistics
one_gene_ct_wald <- TESSERA::calc_Wald_statistics(TESSERA_out, celltype_contrast_mat)
# Remove the raw contrast string for a cleaner display
one_gene_ct_wald <- dplyr::select(one_gene_ct_wald, -c("contrast_string"))
# Preview the first few rows
head(one_gene_ct_wald, 2)
```

### DE within a cell type between conditions {#de-within-celltype}

Beyond comparing cell types, we can construct a contrast matrix to test for differential expression between experimental conditions *within* a specific cell type. For a given cell type $c$, the contrasts of interest are of the following form:
$$
\beta_{\textit{c, DKD}} - \beta_{\textit{c, Control}}, \qquad 
\beta_{\textit{c, HKD}} - \beta_{\textit{c, Control}}, \qquad
\beta_{\textit{c, HKD}} - \beta_{\textit{c, DKD}}.
$$

First, we identify the unique conditions and generate all possible pairwise comparisons.
```{r within_contrast_pairwise}
cell_type_list <- levels(SummarizedExperiment::colData(spe)[, celltype_col])
condition_list <- levels(SummarizedExperiment::colData(spe)[, group_col])

# Generate all pairs of conditions (e.g., Control-DKD, Control-HKD, etc.)
condition_pairs <- combn(condition_list, 2)
# Column names in design matrix
covariate_names <- colnames(X_mat)
```

We iterate through each cell type and, within each, loop through the condition pairs. This identifies the specific entries in the fixed-effect vector $\boldsymbol{\beta}$ that correspond to the intersection of that cell type and those conditions.
```{r within_mat}
# Initialize an empty list to store individual contrast rows
contrast_list <- list()

# Loop over cell types and condition pairs
for (ct in cell_type_list) {
  for (d_idx in 1:ncol(condition_pairs)) {
    # Match the specific column names in the design matrix
    idx1 <- which(covariate_names == paste0(group_col, condition_pairs[1, d_idx], ":", celltype_col, ct))
    idx2 <- which(covariate_names == paste0(group_col, condition_pairs[2, d_idx], ":", celltype_col, ct))
    
    # Skip the contrast if data is missing for either condition in this cell type
    if ((0 == length(idx1)) || (0 == length(idx2))) {
      next
    }
    
    # Create the contrast row
    tmp <- matrix(data = 0, nrow = 1, ncol = length(covariate_names))
    tmp[1, idx1] <- 1
    tmp[1, idx2] <- -1
    
    # Label the row as "Celltype:Condition1-Condition2"
    rownames(tmp) <- c(paste0(ct, ":", condition_pairs[1, d_idx], "-", condition_pairs[2, d_idx]))
    contrast_list[[1 + length(contrast_list)]] <- tmp
  }
}
```

Finally, we bind the individual rows together to create the full contrast matrix.
```{r within_bind}
# Combine the list of rows into a single contrast matrix
contrast_mat <- Reduce(rbind, contrast_list)
colnames(contrast_mat) <- covariate_names

# Preview the first few rows and columns
head(contrast_mat[, 1:5], 2)
```

We now view the results of the between-conditions within-cell type contrasts for the gene *ACTA2* below.
```{r within_wald_stats}
# Compute Wald statistics
one_gene_wald <- TESSERA::calc_Wald_statistics(TESSERA_out, contrast_mat)
# Remove the raw contrast string for a cleaner display
one_gene_wald <- dplyr::select(one_gene_wald, -c("contrast_string"))
# Preview the first few rows
head(one_gene_wald, 2)
```


We can inspect the data frame created in the aggregation script, from running `TESSERA` on the top 3,000 genes.
```{r within_wald_df}
# Filter and preview the aggregated multi-gene results
wald_df <- wald_df[!is.na(wald_df$wald_stat_t), ]
head(wald_df, 2)
```

## Empirical Null Estimation

For a generalized linear mixed model (GLMM), such as the one implemented in `TESSERA`, the null distribution of Wald test statistics is typically not known. 
With a sufficient number of samples (meaning independent experimental units, not individual cells), it will approach a normal distribution or a $\chi_1^2$ distribution (depending on if the statistics are squared). 
However, in the finite-sample regime, GLMMs are prone to bias in both the coefficients and their associated standard errors.
Even with many samples, this bias can shift the null distribution away from theoretical expectations, necessitating an empirical estimation approach.

If we are testing hundreds or thousands of genes, there is a path forward: we can use Efron's idea of estimating the null distribution directly from the data [@efron2004large]. 
In our case, we believe that the square of the Wald statistics computed previously can be modeled as a scaled, non-central $\chi_1^2$ distribution. 
The functions below provide functionality to find the optimal parameter estimates. 
Essentially, we look for a threshold below which most of the statistics come from the null distribution; for these statistics, the $p$-values should look uniformly distributed on $(0, 1)$. 

In general, empirical null estimation procedures rely on having sufficient statistics that come from the null distribution, as otherwise, there will not be enough data to reliably estimate a distribution, or, the statistics used to estimate the distribution will not be from the null distribution. 
Moreover, having more than one contrast (of a similar nature, e.g., DE within a cell type) strengthens the procedure by providing more data and capturing a wider range of variability.

The analysis in this section, and the resulting `wald_df`, focuses exclusively on the between-condition comparisons within each cell type. 
The empirical null distribution described here is derived solely from these specific contrasts.

For this vignette, we have already performed the thresholding procedure using these between-condition within-cell type contrasts. 
We load the resulting object below to demonstrate the downstream analysis.

```{r load_wald_thresh}
# Output of applying TESSERA's empirical null thresholding procedure
wald_thresh_selection <- get_my_data("TESSERA_wald_threshold_out.rds")
```

To execute the full procedure, the following code can be used by setting `eval = TRUE` in the code chunk header. 
Please note that the estimated runtime is approximately 12 minutes on an M1 MacBook with 16 GB of RAM.

```{r empirical_null, eval=FALSE}
t0 <- Sys.time()
# Empirical Null Estimation
# Sweep over thresholds to find the optimal distribution parameters for the empirical null
wald_thresh_selection <- TESSERA::select_Wald_threshold(
  wald_stats = wald_df$wald_stat_t^2,
  quantile_spacing = 0.01,
  metric = "Raw_MSE"
)
t1 <- Sys.time()
cat("Threshold selection of", nrow(wald_df), "statistics took:", 
    round(difftime(t1, t0, units = "min"), 2), "minutes.\n")
```

We can inspect the estimated scale and shift parameters for the fitted scaled, non-central $\chi_1^2$ null distribution.
```{r emp_null_params}
# View the estimated parameters for the scaled, non-central chi-squared distribution
wald_thresh_selection$chi2_params
```

Next, we view the selected Wald statistic threshold.
```{r emp_null_thresh}
# View the selected Wald statistic threshold
wald_thresh_selection$threshold
```

## Obtaining $p$-values

Once the empirical null distribution is fit, we can calculate $p$-values and perform testing for differential expression, in this case, between conditions within a cell type. 

### `TESSERA` $p$-values

We compute the $p$-values and apply the Benjamini-Hochberg (BH) adjustment within each contrast to control the False Discovery Rate (FDR) at a level of 0.05.

```{r sig_genes}
# FDR level
PVAL_THRESH <- 0.05

# Compute p-values using the estimated empirical null distribution
wald_df$wald_pval <- TESSERA::calc_scaled_noncentral_chi2_pvalues(wald_stats = wald_df$wald_stat_t^2,
                                                                  chi2_params = wald_thresh_selection$chi2_params)
# Adjust p-values for multiple testing per contrast
for (cd in unique(wald_df$contrast_description)) {
  wald_df$wald_pval_adj[wald_df$contrast_description == cd] <- p.adjust(wald_df$wald_pval[wald_df$contrast_description == cd], method = "BH")
}
```

We summarize the results by looking at the number of significant genes for each comparison. 
As a sanity check, we see that cell types with very few cells in the dataset, such as RBC (red blood cells), find the fewest number of DE genes. 
```{r n_sig_plot, out.width="100%", fig.height=8, fig.width=10}
n_sig_df <- data.frame(wald_df 
                       %>% dplyr::group_by(contrast_description) 
                       %>% dplyr::summarise(n_sig = sum(wald_pval_adj < PVAL_THRESH, na.rm = TRUE)))
n_sig_df$celltype <- sapply(n_sig_df$contrast_description, function (x) {strsplit(x, ":")[[1]][1]})
n_sig_df$comparison <- sapply(n_sig_df$contrast_description, function (x) {strsplit(x, ":")[[1]][2]})

p <- ggplot2::ggplot(data = n_sig_df, mapping = ggplot2::aes(x = celltype, y = n_sig, fill = comparison))
p <- p + ggplot2::geom_col(position = ggplot2::position_dodge(preserve = "single"))
p <- p + ggplot2::scale_fill_manual(values = c("cornflowerblue", "goldenrod", "coral"))
p <- p + ggplot2::labs(x = "Cell type", 
                       y = "Number of Significant Genes",
                       fill = "Comparison")
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(axis.text.x = ggplot2::element_text(angle = -90, hjust = 0))
p <- p + ggplot2::theme(text = ggplot2::element_text(size = 15))
p
```

### `fdrtool` $p$-values

Another empirical null estimation procedure to obtain $p$-values is the False Non-Discovery Rate (FNDR) method from the `r CRANpkg("fdrtool")` package [@strimmer2008fdrtool]. 
While both empirical null estimation procedures can be used to generate $p$-values, there are key trade-offs to consider: `fdrtool` offers a faster runtime than `TESSERA`'s thresholding procedure, but `TESSERA` makes fewer underlying assumptions about the data distribution. 
Both methods achieve highly comparable performance on this class of spatial transcriptomics data as can be seen in the `TESSERA` methodological manuscript [@constantine2026tessera].
```{r fdrtool_sig_genes}
# Estimate p-values using fdrtool's FNDR method
fndr_out <- fdrtool::fdrtool(wald_df$wald_stat_t,
                             cutoff.method = "fndr",
                             plot = FALSE)


# Store fdrtool p-values
wald_df$wald_pval_fndr <- fndr_out$pval

# Adjust fdrtool p-values for multiple testing
wald_df$wald_pval_adj_fndr <- NA
for (cd in unique(wald_df$contrast_description)) {
  wald_df$wald_pval_adj_fndr[wald_df$contrast_description == cd] <- p.adjust(wald_df$wald_pval_fndr[wald_df$contrast_description == cd], method = "BH")
}
```

# Session information

```{r sessioninfo}
devtools::session_info()
```

# References
